---
title: "EM for Mixtures of t-Distributions"
subtitle: "Computational Statistics — Assignment 3, topic B"
format:
  revealjs:
    theme: default
    css: styles.css
    slide-number: true
    smaller: true
    scrollable: true
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
    width: 1200
    height: 750
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    fig.width: 10
    fig.height: 4.6
    fig.align: center
    comment: "#>"
---

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(patchwork)
library(testthat)
source("R/load-v1.R")
theme_set(theme_minimal(base_size = 15))
options(digits = 5, width = 110)

# print the source of a function (with comments) in a code chunk
show_fun <- function(name) {
  paste0(name, " <- ", paste(as.character(attr(get(name), "srcref")), collapse = "\n"))
}
```

## Roadmap

1. **A simple implementation** — the formulas coded directly, on the density scale, with R's `dt()`
2. **Is it correct?** — checks against theory, numerical derivatives, simulation and R packages (`teigen`, `mclust`)
3. **Where is it slow?** — profiling with `Rprof` and `profvis`
4. **A faster implementation** — a log-scale version, and an S3 design with two classes so the two implementations share the same GEM code and tests

Throughout: [boxes like this one]{.lesson} mark what I learned along the way.

# A simple implementation

## Two-component mixture of $t$-distributions

$$
f(x \mid \theta) = p\, f(x \mid \mu_1, \sigma_1^2, \nu_1) + (1 - p)\, f(x \mid \mu_2, \sigma_2^2, \nu_2),
\qquad \theta = (p, \mu_1, \mu_2, \sigma_1, \sigma_2)
$$

with $f(x \mid \mu, \sigma^2, \nu) = \frac{\Gamma((\nu+1)/2)}{\sqrt{\pi\nu\sigma^2}\,\Gamma(\nu/2)}\left(1 + \frac{(x-\mu)^2}{\nu\sigma^2}\right)^{-(\nu+1)/2}$ and $\nu_1, \nu_2$ fixed.

:::: {.columns}
::: {.column width="45%"}
Simulated data as in the assignment:

```{r}
#| label: sim-data
set.seed(42)
nu <- c(5, 10)
par_true <- c(p = 0.6, mu1 = 0, mu2 = 3,
              sigma1 = 1, sigma2 = 0.7)
x <- rtmix(400, par_true, nu)
par0 <- default_start(x)
round(par0, 3)
```

`default_start()`: $p = 0.5$, quartiles for $\mu$, $\text{sd}(x)/2$ for $\sigma$.
:::

::: {.column width="55%"}
```{r}
#| label: fig-data
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.3
grid <- seq(min(x), max(x), length.out = 400)
ggplot(data.frame(x), aes(x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "grey85", colour = "white") +
  geom_line(data = data.frame(x = grid, d = dtmix(grid, par_true, nu)), aes(x, d), linewidth = 1) +
  labs(y = "density", title = "N = 400, true density")
```
:::
::::

## Missing data: the component labels

**Idea:** if we knew which component each $x_i$ came from, estimation would be easy — just like the two coins in lecture 9.

Introduce the (unobserved) label $Z_i \in \{1, 2\}$:
$$
P(Z_i = 1) = p, \qquad X_i \mid Z_i = j \sim t(\mu_j, \sigma_j^2, \nu_j).
$$

Write $f_j(x) = f(x \mid \mu_j, \sigma_j^2, \nu_j)$. The joint density of one complete observation $(x_i, z_i)$ is
$$
f(x_i, z_i \mid \theta) = \big(p\, f_1(x_i)\big)^{\mathbf{1}(z_i = 1)} \big((1 - p)\, f_2(x_i)\big)^{\mathbf{1}(z_i = 2)}.
$$

. . .

**Check:** summing over $z_i \in \{1, 2\}$ gives $p f_1(x_i) + (1-p) f_2(x_i)$ — exactly the mixture density. So $(X, Z)$ is a valid complete-data model for our observed $X$.

**Complete-data log-likelihood** (what we could maximize if $Z$ were observed):
$$
\ell_c(\theta) = \sum_{i=1}^N \mathbf{1}(Z_i = 1)\big(\log p + \log f_1(x_i)\big) + \mathbf{1}(Z_i = 2)\big(\log(1-p) + \log f_2(x_i)\big)
$$

::: notes
The trick of EM is to choose missing data such that the complete-data problem is easy. Here the missing data are the labels. The check that summing out z gives the mixture density is what makes this a legal choice of complete data.
:::

## E-step: from $\ell_c$ to $Q$

$Q(\theta \mid \theta') = E_{\theta'}\big(\ell_c(\theta) \mid X = x\big)$ — average $\ell_c$ over what the labels could be, given the data and the current guess $\theta'$.

**Step 1 — linearity.** Given $x$, the terms $\log p + \log f_1(x_i)$ are just numbers, so the expectation only hits the indicators:
$$
E_{\theta'}\big(\mathbf{1}(Z_i = 1) \mid X = x\big) = P_{\theta'}(Z_i = 1 \mid X_i = x_i) =: \gamma_i
$$

**Step 2 — Bayes' formula.**
$$
\gamma_i = \frac{P(Z_i = 1)\, f(x_i \mid Z_i = 1)}{f(x_i)} = \frac{p' f_1'(x_i)}{p' f_1'(x_i) + (1 - p') f_2'(x_i)}
$$

**Result:**
$$
Q(\theta \mid \theta') = \sum_{i=1}^N \gamma_i \big(\log p + \log f_1(x_i)\big) + (1 - \gamma_i)\big(\log(1-p) + \log f_2(x_i)\big)
$$

The $\gamma_i$ are computed from the **old** $\theta'$ and are fixed numbers ("soft labels"); $\theta$ in the logs is the **new** parameter we optimize over.

```{r}
#| label: fig-gamma
#| echo: false
#| fig-height: 2.6
par_hat <- gem(x, nu, eps = 1e-14)
gdf <- rbind(
  data.frame(x = grid, gamma = e_step(par0, grid, nu), theta = "at the start value"),
  data.frame(x = grid, gamma = e_step(par_hat, grid, nu), theta = "at the estimate")
)
ggplot(gdf, aes(x, gamma, colour = theta)) +
  geom_line(linewidth = 1) +
  geom_rug(data = data.frame(x), aes(x), inherit.aes = FALSE, alpha = 0.3) +
  labs(y = expression(gamma == P(Z == 1 ~ "|" ~ X == x)), colour = NULL)
```

## M-step: maximize $Q$ — what can be solved?

$Q$ splits into three parts that can be maximized separately:
$$
Q = \underbrace{\sum_i \gamma_i \log p + (1 - \gamma_i)\log(1-p)}_{\text{only } p}
  + \underbrace{\sum_i \gamma_i \log f_1(x_i)}_{\text{only } \mu_1, \sigma_1}
  + \underbrace{\sum_i (1 - \gamma_i) \log f_2(x_i)}_{\text{only } \mu_2, \sigma_2}
$$

**$p$:** $\ \partial_p Q = \frac{\sum_i \gamma_i}{p} - \frac{N - \sum_i \gamma_i}{1 - p} = 0 \;\Rightarrow\; p = \frac{1}{N}\sum_i \gamma_i$ — closed form.

. . .

**$(\mu_1, \sigma_1)$:** a *weighted* $t$ log-likelihood. With $r = x - \mu_1$ and $\log f_1(x) = c(\nu_1) - \log \sigma_1 - \frac{\nu_1 + 1}{2}\log\big(1 + \frac{r^2}{\nu_1 \sigma_1^2}\big)$, the chain rule gives
$$
\partial_{\mu_1} \log f_1(x) = \frac{w\, r}{\sigma_1^2},
\qquad
\partial_{\sigma_1} \log f_1(x) = -\frac{1}{\sigma_1} + \frac{w\, r^2}{\sigma_1^3},
\qquad w = \frac{\nu_1 + 1}{\nu_1 + r^2/\sigma_1^2},
$$
so
$$
\partial_{\mu_1} Q = \sum_i \gamma_i \frac{w_{i1} r_{i1}}{\sigma_1^2}, \qquad
\partial_{\sigma_1} Q = \sum_i \frac{\gamma_i}{\sigma_1}\left(\frac{w_{i1} r_{i1}^2}{\sigma_1^2} - 1\right) \qquad (\text{component 2: } 1 - \gamma_i).
$$

. . .

Setting $\partial_{\mu_1} Q = 0$ gives $\mu_1 = \frac{\sum_i \gamma_i w_{i1} x_i}{\sum_i \gamma_i w_{i1}}$ — but $w_{i1}$ depends on $\mu_1$ and $\sigma_1$ themselves: a fixed-point equation, not a formula.

[No closed-form M-step for $(\mu, \sigma)$. **Generalized EM:** we only need $Q(\theta_{n+1} \mid \theta_n) \ge Q(\theta_n \mid \theta_n)$, so: exact update of $p$ + one uphill (gradient) step for $(\mu_1, \mu_2, \sigma_1, \sigma_2)$.]{.problem}

## The code: density, log-likelihood and E-step

Written exactly as the formulas, directly on the density scale. The location-scale $t$-density is $f(x \mid \mu, \sigma^2, \nu) = \frac{1}{\sigma} f_{t_\nu}\big(\frac{x - \mu}{\sigma}\big)$, so R's `dt()` does the work:

::: {.small-code}
```{r}
#| eval: false
#| code: !expr c(show_fun("dt_ls"), "", show_fun("dtmix"), "", show_fun("loglik"), "", show_fun("e_step"))
```
:::

## The code: $Q$ and its gradient

`Q()` and `grad_Q()` take the responsibilities `gam` instead of $\theta'$: the $\gamma_i$ are *all* that $Q$ needs from $\theta'$, and the E-step is then done once per iteration.

| Math | Code |
|------|------|
| $\gamma_i \big(\log p + \log f_1(x_i)\big)$ | `gam * log(par[1] * dt_ls(x, par[2], par[4], nu[1]))` |
| $r_{i1}$, $w_{i1}$ | `r1 <- x - par[2]`, `w1 <- (nu[1] + 1) / (nu[1] + r1^2 / s1^2)` |
| $\partial_{\mu_1} Q = \sum_i \gamma_i w_{i1} r_{i1} / \sigma_1^2$ | `sum(gam * w1 * r1) / s1^2` |

::: {.small-code}
```{r}
#| eval: false
#| code: !expr c(show_fun("Q"), "", show_fun("grad_Q"))
```
:::

## One gradient step — but how long?

First iteration from `par0`: E-step, update $p$, and step along the gradient with a few *fixed* step sizes:

```{r}
#| label: fixed-steps
gam <- e_step(par0, x, nu)
par1 <- par0
par1[1] <- mean(gam)                       # exact M-step for p
g <- grad_Q(par1, gam, x, nu)
g[1] <- 0
step_sizes <- c(0.1, 0.01, 0.001, 0.0001)
new_par <- t(sapply(step_sizes, function(s) par1 + s * g))
data.frame(step = step_sizes,
           new_sigma1 = new_par[, 4], new_sigma2 = new_par[, 5],
           change_in_Q = apply(new_par, 1, function(p) Q(p, gam, x, nu) - Q(par1, gam, x, nu)))
```

[The gradient only gives the *direction*. A very long step gives an invalid $\sigma < 0$ (so `NaN`); a long step jumps past the top and $Q$ **decreases** (then it is not a GEM step); too short a step barely moves. The right length depends on the data: the gradient is a sum over $N$ points and scales like $N / \sigma^2$.]{.problem}

. . .

**Solution — backtracking:** start with a long step and halve it until $Q$ has increased enough (Armijo condition $Q(\theta + s g) \ge Q(\theta) + c\, s\, \|g\|^2$).

## The code: the GEM step

::: {.small-code}
```{r}
#| eval: false
#| code: !expr show_fun("create_gem_step")
```
:::

`create_gem_step()` is a **function factory**: it stores the data and returns a function that maps $\theta_n \mapsto \theta_{n+1}$.

## The code: the driver

::: {.small-code}
```{r}
#| eval: false
#| code: !expr c(show_fun("em"), "", show_fun("gem"))
```
:::

`em()` is the driver from the lecture, with the same stopping rule $\|\theta_n - \theta_{n-1}\|^2 \le \varepsilon(\|\theta_n\|^2 + \varepsilon)$.

## Running it

:::: {.columns}
::: {.column width="45%"}
```{r}
#| label: run-gem
fit <- gem(x, nu)
fit
loglik(fit, x, nu)
```
:::
::: {.column width="55%"}
```{r}
#| label: fig-fit
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.3
p <- fit
df_fit <- data.frame(
  x = rep(grid, 3),
  density = c(dtmix(grid, p, nu), p[1] * dt_ls(grid, p[2], p[4], nu[1]), (1 - p[1]) * dt_ls(grid, p[3], p[5], nu[2])),
  curve = rep(c("mixture", "component 1", "component 2"), each = length(grid))
)
ggplot(data.frame(x), aes(x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "grey85", colour = "white") +
  geom_line(data = df_fit, aes(x, density, colour = curve, linetype = curve), linewidth = 0.9) +
  scale_linetype_manual(values = c(2, 2, 1)) +
  labs(colour = NULL, linetype = NULL, y = "density", title = "Fitted mixture")
```
:::
::::

# Is it correct?

## How can we test that the implementation is correct?

Every part is checked against something **independent** of the code being tested:

| What | Independent reference |
|------|-----------------------|
| $t$-density `dt_ls()` | the formula from the assignment, written with `gamma()` |
| mixture density | integrates to 1 (`integrate()`) |
| simulation `rtmix()` | Kolmogorov–Smirnov test against the mixture CDF (`pt()`) |
| E-step | calibration on simulated data with **known** labels |
| `grad_Q()` | numerical differentiation (`numDeriv::grad()`) |
| Fisher's identity $\nabla\ell(\theta) = \nabla_\theta Q(\theta \mid \theta)$ | `numDeriv::grad(loglik)` |
| the GEM property | $Q$ and $\ell$ increase in every iteration |
| the maximum | `optim()` (BFGS) on the log-likelihood |
| the whole fit | R package **teigen** ($t$-mixtures, common $\nu$) |
| the whole fit | R package **mclust** (Gaussian mixtures = the limit $\nu \to \infty$) |
| statistical sanity | 100 simulated data sets: estimates centred at the truth |

All of it is collected in `tests/test-simple.R` (**testthat**).

## Building blocks

```{r}
#| label: test-blocks
# 1. t-density vs. the formula from the assignment (no dt() involved)
f_formula <- function(x, mu, sigma, nu) {
  gamma((nu + 1) / 2) / (sqrt(pi * nu * sigma^2) * gamma(nu / 2)) *
    (1 + (x - mu)^2 / (nu * sigma^2))^(-(nu + 1) / 2)
}
z <- seq(-10, 10, length.out = 101)
all.equal(dt_ls(z, 1, 2, 3), f_formula(z, 1, 2, 3))

# 2. the mixture density integrates to 1
integrate(dtmix, -Inf, Inf, par = par_true, nu = nu)

# 3. simulated data have the right distribution
ptmix <- function(q, par, nu) {
  par[1] * pt((q - par[2]) / par[4], nu[1]) + (1 - par[1]) * pt((q - par[3]) / par[5], nu[2])
}
set.seed(3)
ks.test(rtmix(20000, par_true, nu), ptmix, par = par_true, nu = nu)$p.value
```

## Is the E-step right? A calibration check

If $\gamma_i = P(Z_i = 1 \mid x_i)$ is computed correctly, then among all points with $\gamma \approx 0.7$, about 70% must come from component 1. Simulate $N = 20\,000$ points **with** their labels and check:

```{r}
#| label: calibration
set.seed(3)
d <- sim_tmix(20000, par_true, nu)        # data frame with x and the true label z
d$gamma <- e_step(par_true, d$x, nu)
d$bin <- cut(d$gamma, seq(0, 1, 0.1), include.lowest = TRUE)
calib <- data.frame(
  mean_gamma = tapply(d$gamma, d$bin, mean),
  frac_comp1 = tapply(d$z == 1, d$bin, mean),
  n = tapply(d$gamma, d$bin, length)
)
```

```{r}
#| label: fig-calibration
#| echo: false
#| fig-height: 3.3
calib$se <- sqrt(calib$mean_gamma * (1 - calib$mean_gamma) / calib$n)
ggplot(calib, aes(mean_gamma, frac_comp1)) +
  geom_abline(linetype = 2, colour = "grey50") +
  geom_errorbar(aes(ymin = frac_comp1 - 1.96 * se, ymax = frac_comp1 + 1.96 * se), width = 0.015) +
  geom_point(size = 2.5) +
  coord_equal() +
  labs(x = expression("average " * gamma * " in bin"), y = "fraction from component 1")
```

## Derivatives: analytic vs. numerical

```{r}
#| label: test-derivatives
par_test <- c(0.4, -0.5, 2.5, 1.3, 0.9)
gam0 <- e_step(par0, x, nu)

# analytic gradient of Q vs. numerical differentiation
g <- rbind(analytic  = grad_Q(par_test, gam0, x, nu),
           numerical = numDeriv::grad(function(p) Q(p, gam0, x, nu), par_test))
colnames(g) <- names(par_true)
g

# Fisher's identity: grad_Q(theta | theta) equals the gradient of the log-likelihood
g <- rbind(fisher_identity = grad_Q(par_test, e_step(par_test, x, nu), x, nu),
           numerical       = numDeriv::grad(function(p) loglik(p, x, nu), par_test))
colnames(g) <- names(par_true)
g
```

[Deriving by hand and checking against `numDeriv` catches sign and chain-rule errors before they turn into a mysteriously slow or wrong algorithm.]{.lesson}

## Does it behave like a GEM?

```{r}
#| label: test-monotone
step <- create_gem_step(x, nu)
par <- par0
trace <- NULL
for (n in 1:40) {
  gam <- e_step(par, x, nu)
  par_new <- step(par)
  trace <- rbind(trace, data.frame(
    n = n,
    Q_increase = Q(par_new, gam, x, nu) - Q(par, gam, x, nu),
    loglik = loglik(par_new, x, nu)
  ))
  par <- par_new
}
c(smallest_Q_increase = min(trace$Q_increase), smallest_loglik_increase = min(diff(trace$loglik)))
```

```{r}
#| label: fig-monotone
#| echo: false
#| fig-height: 2.8
ggplot(trace, aes(n, pmax(loglik(par_hat, x, nu) - loglik, 1e-14))) +
  geom_line(linewidth = 0.9) + geom_point() + scale_y_log10() +
  labs(x = "iteration", y = expression(hat(l) - l(theta[n])))
```

$Q$ increases in every step (guaranteed by the backtracking), and so does $\ell$ — as the EM theory promises. Convergence is linear (a straight line on log scale).

## Same maximum as `optim()`

Maximize the observed log-likelihood directly with BFGS, in the unconstrained parametrization $(\operatorname{logit} p, \mu_1, \mu_2, \log\sigma_1, \log\sigma_2)$:

```{r}
#| label: test-optim
nll <- function(eta) -loglik(c(plogis(eta[1]), eta[2:3], exp(eta[4:5])), x, nu)
eta <- optim(c(qlogis(par0[1]), par0[2:3], log(par0[4:5])), nll,
             method = "BFGS", control = list(reltol = 1e-14, maxit = 1000))$par
par_optim <- c(plogis(eta[1]), eta[2:3], exp(eta[4:5]))

rbind(gem = par_hat, optim = par_optim)
c(gem = loglik(par_hat, x, nu), optim = loglik(par_optim, x, nu))
```

## Against an R package: teigen

**teigen** fits mixtures of $t$-distributions with its own EM-type algorithm. It requires a *common* $\nu$, so we simulate with $\nu_1 = \nu_2 = 5$ and hold $\nu$ fixed (`dfupdate = FALSE`):

```{r}
#| label: test-teigen
set.seed(1)
x5 <- rtmix(1000, par_true, c(5, 5))
par_gem5 <- gem(x5, c(5, 5), eps = 1e-14)

tg <- teigen::teigen(x5, Gs = 2, models = "univUU", dfstart = 5, dfupdate = FALSE,
                     scale = FALSE, verbose = FALSE, eps = c(1e-12, 1e-12))
o <- order(tg$parameters$mean)              # teigen may order the components differently
par_teigen <- c(tg$parameters$pig[o][1], tg$parameters$mean[o],
                sqrt(tg$parameters$sigma[1, 1, o]))   # teigen reports variances

rbind(gem = par_gem5, teigen = par_teigen)
c(gem = loglik(par_gem5, x5, c(5, 5)), teigen = tg$logl)
```

## Against an R package: mclust (the Gaussian limit)

As $\nu \to \infty$ the $t$-distribution becomes the normal distribution, so with $\nu = 10^6$ our GEM must agree with a Gaussian mixture fitted by **mclust**:

```{r}
#| label: test-mclust
suppressPackageStartupMessages(library(mclust))
set.seed(2)
xg <- c(rnorm(600, 0, 1), rnorm(400, 3, 0.7))
par_gem_gauss <- gem(xg, c(1e6, 1e6), eps = 1e-16, maxit = 5000)

mclust_par <- function(tol) {
  mc <- Mclust(xg, G = 2, modelNames = "V", verbose = FALSE,
               control = emControl(tol = c(tol, sqrt(.Machine$double.eps))))
  unname(c(mc$parameters$pro[1], mc$parameters$mean, sqrt(mc$parameters$variance$sigmasq)))
}
rbind(gem = par_gem_gauss,
      mclust_default_tol = mclust_par(1e-5),
      mclust_tol_1e12 = mclust_par(1e-12))
```

[At first mclust and my GEM disagreed in the second decimal — and my log-likelihood was *higher*. mclust's default tolerance (`1e-5` relative change) stops early. With the same strict tolerance the two agree. A fair comparison needs the same convergence criterion.]{.lesson}

## Statistical sanity: 100 simulated data sets

```{r}
#| label: test-sim
set.seed(123)
est <- t(replicate(100, gem(rtmix(400, par_true, nu), nu)))
data.frame(true = par_true, mean_estimate = colMeans(est), sd = apply(est, 2, sd))
```

The estimates are centred at the true values (up to Monte Carlo error $\approx \text{sd}/\sqrt{100}$).

## The test suite

```{r}
#| label: run-tests
test_res <- testthat::test_file("tests/test-simple.R", reporter = "silent")
res <- as.data.frame(test_res)[, c("test", "nb", "failed", "error")]
res
```

**`r sum(res$nb)` expectations, `r sum(res$failed)` failures.**

# Where is it slow?

## Setup: a larger data set

$N = 5000$ observations, so the time is spent on computation and not on overhead:

```{r}
#| label: prof-setup
set.seed(7)
x_big <- rtmix(5000, par_true, nu)
fit_big <- gem(x_big, nu)
n_iter <- attr(fit_big, "iter")
n_iter

bm_fit <- bench::mark(gem(x_big, nu), iterations = 5)
bm_fit[, c("median", "mem_alloc")]
```

One fit takes about **`r format(bm_fit$median)`** and `r n_iter` iterations. Where does the time go?

## Profiling by function

`Rprof` samples the call stack every 5 ms (here over 5 fits):

```{r}
#| label: prof-functions
prof_file <- profile_expr(for (i in 1:5) gem(x_big, nu))
prof <- summaryRprof(prof_file)
our_funs <- dQuote(c("em_step", "e_step", "Q", "grad_Q", "dt_ls", "dt", "log", "sum"), FALSE)
prof$by.total[rownames(prof$by.total) %in% our_funs, ]
```

`total.pct`: share of samples where the function is *on* the stack. `self.pct`: share where it is the function *actually running*.

. . .

- `Q()` is on the stack in about `r round(prof$by.total['"Q"', "total.pct"])`% of the time — but it is only called inside the GEM step.
- `dt()` itself is running in about `r round(prof$by.self['"dt"', "self.pct"])`% of the time.

Interactively: `profvis::profvis(gem(x_big, nu))` (see `R/profiling-demo.R`).

## Profiling by line

The same samples, attributed to source lines (`Rprof(line.profiling = TRUE)`):

```{r}
#| label: prof-lines
profile_lines(prof_file)
```

The line in `gem.R` that dominates is the call to `Q()` **inside the backtracking loop**. Inside `Q()` and `e_step()`, the time is spent in `dt_ls()`, i.e. in `dt()`.

## Why? Number of calls × cost per call

```{r}
#| label: prof-calls
calls <- count_calls(gem(x_big, nu), c("e_step", "Q", "grad_Q", "dt_ls"))
calls

gam_big <- e_step(fit_big, x_big, nu)
cost <- bench::mark(
  e_step = e_step(fit_big, x_big, nu),
  Q      = Q(fit_big, gam_big, x_big, nu),
  grad_Q = grad_Q(fit_big, gam_big, x_big, nu),
  dt_ls  = dt_ls(x_big, 0, 1, 5),
  check = FALSE
)
per_call <- setNames(as.numeric(cost$median), as.character(cost$expression))
data.frame(
  calls_per_fit = calls,
  calls_per_iteration = round(calls / n_iter, 1),
  ms_per_call = round(1000 * per_call[names(calls)], 3),
  total_ms = round(1000 * calls * per_call[names(calls)]),
  share_of_fit = paste0(round(100 * calls * per_call[names(calls)] / as.numeric(bm_fit$median)), "%")
)
```

(`dt_ls()` is called *inside* `e_step()` and `Q()`, so its row overlaps with theirs: 2 calls per E-step + 2 per $Q$.)

## Conclusion: three hotspots

1. **The line search.** $Q$ is evaluated `r round(calls[["Q"]] / n_iter, 1)` times per iteration (`r calls[["Q"]]` calls in `r n_iter` iterations), because the backtracking restarts from step size 1 in *every* iteration and needs about `r round(calls[["Q"]] / n_iter - 2)` halvings before the step is short enough ($\approx 2^{-`r round(calls[["Q"]] / n_iter - 2)`}$).
2. **`dt()`.** Most of the time is spent inside R's `dt()`. Every $Q$ and every E-step evaluates the $t$-density twice for all $N$ points: `r calls[["dt_ls"]]` density evaluations of length `r length(x_big)` per fit.
3. **Repeated work.** The E-step computes $p f_1(x_i)$ and $(1-p) f_2(x_i)$; right after, $Q$ (at the same parameters, as `Q_old`) computes the same densities again, and `grad_Q()` recomputes the residuals.

[The profile is dominated by one line (the $Q$ call in the line search) and one primitive (`dt()`). Counting calls and timing each building block explains *why*: total time ≈ number of calls × cost per call.]{.lesson}

::: notes
Transition: hotspot 2 (dt) is attacked next with a log-scale implementation. Hotspot 1 (the number of Q calls) is still there afterwards, which the new profile at the end shows.
:::

# A faster implementation: the log scale

```{r}
#| label: load-s3
#| include: false
# Version 2 of the code: the building blocks become S3 generics (replaces the plain functions)
source("R/load.R")
```

## From the profile to a plan

**What the profile showed:** `Q()` takes `r round(prof$by.total['"Q"', "total.pct"])`% of the time, and `dt()` alone `r round(prof$by.self['"dt"', "self.pct"])`%.

**Observation:** $Q$ only needs **log**-densities,
$$
Q(\theta \mid \theta') = \sum_i \gamma_i \big(\log p + \log f_1(x_i)\big) + (1 - \gamma_i)\big(\log(1-p) + \log f_2(x_i)\big),
$$
but version 1 computes the density with `dt()` and then takes `log()` of it — a detour through the density scale.

. . .

**Idea:** write the log-density out and compute it directly,
$$
\log f(x) = \underbrace{\log\Gamma\big(\tfrac{\nu+1}{2}\big) - \log\Gamma\big(\tfrac{\nu}{2}\big) - \tfrac12\log(\pi\nu)}_{c(\nu)\text{: a constant, computed once}} - \log\sigma - \tfrac{\nu+1}{2}\log\Big(1 + \tfrac{(x-\mu)^2}{\nu\sigma^2}\Big).
$$
With $a_i = \log p + \log f_1(x_i)$ and $b_i = \log(1-p) + \log f_2(x_i)$:

| Building block | Version 1 (density scale) | Log scale |
|---|---|---|
| $Q$ | `log(p * dt(...))`, twice per point | $\sum_i \gamma_i a_i + (1-\gamma_i) b_i$ — no `dt()`, no `log()` |
| E-step | $\gamma_i = \frac{p f_1}{p f_1 + (1-p) f_2}$ | $\gamma_i = \frac{1}{1 + e^{b_i - a_i}}$ = `plogis(a - b)` |
| log-likelihood | `sum(log(dtmix(...)))` | $\log(e^{a_i} + e^{b_i})$ with the log-sum-exp trick |

. . .

**To compare the two fairly** they must run through the *same* GEM code and the *same* tests. So the code is restructured with S3: one class per implementation, and everything they share in a common parent class.

## The S3 design: two classes with a common parent

:::: {.columns}
::: {.column width="48%"}
```{=html}
<svg viewBox="0 0 520 400" style="width: 100%; max-width: 540px;" xmlns="http://www.w3.org/2000/svg" font-family="monospace" font-size="15" role="img" aria-label="Class diagram: tmix is the parent of tmix_density and tmix_log">
  <g stroke="#2c7fb8" stroke-width="1.5" fill="white">
    <rect x="150" y="8" width="220" height="150"/>
    <rect x="150" y="8" width="220" height="32" fill="#dcebf7"/>
    <line x1="150" y1="92" x2="370" y2="92"/>
    <rect x="8" y="230" width="230" height="162"/>
    <rect x="8" y="230" width="230" height="32" fill="#dcebf7"/>
    <line x1="8" y1="294" x2="238" y2="294"/>
    <rect x="282" y="230" width="230" height="162"/>
    <rect x="282" y="230" width="230" height="32" fill="#dcebf7"/>
    <line x1="282" y1="294" x2="512" y2="294"/>
    <polyline points="123,230 123,200 215,200 215,172" fill="none"/>
    <polygon points="215,158 207,172 223,172"/>
    <polyline points="397,230 397,200 305,200 305,172" fill="none"/>
    <polygon points="305,158 297,172 313,172"/>
  </g>
  <g fill="#222">
    <text x="260" y="30" text-anchor="middle" font-weight="bold">tmix</text>
    <text x="162" y="62">x   (data)</text>
    <text x="162" y="82">nu  (shape parameters)</text>
    <text x="162" y="114">grad_Q()</text>
    <text x="162" y="132">fit()</text>
    <text x="162" y="150">print()</text>
    <text x="123" y="252" text-anchor="middle" font-weight="bold">tmix_density</text>
    <text x="20" y="282" fill="#777">(no extra fields)</text>
    <text x="20" y="316">loglik()</text>
    <text x="20" y="334">e_step()</text>
    <text x="20" y="352">Q()</text>
    <text x="20" y="380" fill="#2c7fb8">density scale: dt()</text>
    <text x="397" y="252" text-anchor="middle" font-weight="bold">tmix_log</text>
    <text x="294" y="282">log_const  (= c(nu))</text>
    <text x="294" y="316">loglik()</text>
    <text x="294" y="334">e_step()</text>
    <text x="294" y="352">Q()</text>
    <text x="294" y="380" fill="#2c7fb8">log scale: log1p()</text>
  </g>
</svg>
```
:::
::: {.column width="52%"}
**`tmix`** (parent) holds what is the same for both:

- the data `x` and the shape parameters `nu`
- `grad_Q()`: the gradient is the same formula
- `fit()`: the GEM algorithm
- `print()`

**`tmix_density`**: version 1 — `loglik()`, `e_step()` and `Q()` with `dt()`.

**`tmix_log`**: the same three methods on log scale, plus the constant $c(\nu)$ stored in the object.

Only what *differs* between the implementations is written twice.
:::
::::

## How an object is created

In S3 a class is not a separate definition: an object is an ordinary **list with a class label**, made by a **constructor** function.

::: {.small-code}
```{r}
#| eval: false
#| code: !expr c(show_fun("new_tmix"), "", show_fun("tmix_density"), "", show_fun("tmix_log"))
```
:::

```{r}
#| label: s3-object
m_log <- tmix_log(x, nu)
class(m_log)
str(unclass(m_log))
m_log
```

The label `c("tmix_log", "tmix")` says: "I am a `tmix_log` — and where nothing more specific exists, a `tmix`". That is the inheritance.

## Generics, methods and dispatch

A **generic** only dispatches; a **method** is an ordinary function named `generic.class`:

```{r}
#| eval: false
#| code: !expr c(show_fun("loglik"), show_fun("e_step"), show_fun("Q"), show_fun("grad_Q"), show_fun("fit"))
```

| Generic | `tmix_density` | `tmix_log` |
|---|---|---|
| `loglik()` | `loglik.tmix_density` | `loglik.tmix_log` |
| `e_step()` | `e_step.tmix_density` | `e_step.tmix_log` |
| `Q()` | `Q.tmix_density` | `Q.tmix_log` |
| `grad_Q()` | `grad_Q.tmix` (inherited) | `grad_Q.tmix` (inherited) |
| `fit()`, `print()` | `fit.tmix`, `print.tmix` (inherited) | `fit.tmix`, `print.tmix` (inherited) |

```{r}
#| label: s3-methods
methods(class = "tmix_log")
methods(class = "tmix")
```

`e_step(m_log, par)`: R looks for `e_step.tmix_log` — found. `grad_Q(m_log, par, gam)`: no `grad_Q.tmix_log`, so R tries the next label and uses `grad_Q.tmix`.

## The density class: version 1 as methods

The code is the same as version 1 — only the data now come from the model object (`model$x`, `model$nu`):

::: {.small-code}
```{r}
#| eval: false
#| code: !expr c(show_fun("loglik.tmix_density"), "", show_fun("e_step.tmix_density"), "", show_fun("Q.tmix_density"))
```
:::

## The log class

::: {.small-code}
```{r}
#| eval: false
#| code: !expr c(show_fun("ldt_ls"), "", show_fun("log_comp"), "", show_fun("loglik.tmix_log"), "", show_fun("e_step.tmix_log"), "", show_fun("Q.tmix_log"))
```
:::

`ldt_ls()` and `log_comp()` are helper functions, not methods. `log_comp()` computes $a_i$ and $b_i$, which all three methods use.

## One GEM algorithm for both classes

The GEM step only talks to the model through the generics:

::: {.small-code}
```{r}
#| eval: false
#| code: !expr c(show_fun("create_gem_step"), "", show_fun("fit.tmix"))
```
:::

The only change from version 1: `e_step(par, x, nu)` becomes `e_step(model, par)`, and so on.

```{r}
#| label: s3-fit
rbind(density = fit(tmix_density(x, nu)),
      log     = fit(tmix_log(x, nu)))
```

## The same tests for both classes

`tests/test-classes.R` loops over the two classes, so *every* check from the testing section runs for both. In addition it checks the two implementations against each other, and that the log scale survives an extreme outlier where the density scale underflows:

```{r}
#| label: s3-outlier
x_out <- c(x, -1e60)
c(density = tail(e_step(tmix_density(x_out, nu), par_true), 1),
  log     = tail(e_step(tmix_log(x_out, nu), par_true), 1))
```

```{r}
#| label: s3-tests
test_res2 <- testthat::test_file("tests/test-classes.R", reporter = "silent")
res2 <- as.data.frame(test_res2)[, c("test", "nb", "failed", "error")]
res2
```

**`r sum(res2$nb)` expectations, `r sum(res2$failed)` failures.**

## Is it faster?

The same data as in the profiling ($N = 5000$). `check` makes `bench::mark` verify that both classes give the same estimate:

```{r}
#| label: s3-bench
model_density <- tmix_density(x_big, nu)
model_log <- tmix_log(x_big, nu)
same_fit <- function(a, b) isTRUE(all.equal(as.vector(a), as.vector(b), tolerance = 1e-8))
bm_classes <- bench::mark(density = fit(model_density), log = fit(model_log),
                          check = same_fit, iterations = 10)
bm_classes[, c("expression", "median", "mem_alloc")]
```

```{r}
#| label: s3-blocks
par_b <- fit(model_log)
gam_b <- e_step(model_log, par_b)
blocks <- bench::mark(
  Q_density = Q(model_density, par_b, gam_b),           Q_log = Q(model_log, par_b, gam_b),
  e_step_density = e_step(model_density, par_b),        e_step_log = e_step(model_log, par_b),
  loglik_density = loglik(model_density, par_b),        loglik_log = loglik(model_log, par_b),
  check = FALSE
)
med <- setNames(1e6 * as.numeric(blocks$median), as.character(blocks$expression))
data.frame(
  density_us = round(med[c("Q_density", "e_step_density", "loglik_density")]),
  log_us     = round(med[c("Q_log", "e_step_log", "loglik_log")]),
  speed_up   = round(med[c("Q_density", "e_step_density", "loglik_density")] /
                     med[c("Q_log", "e_step_log", "loglik_log")], 1),
  row.names  = c("Q", "e_step", "loglik")
)
```

The whole fit is **`r round(as.numeric(bm_classes$median[1] / bm_classes$median[2]), 1)` times faster** on log scale, with the same estimate and the same number of iterations — because $Q$, which dominated the profile, got the largest speed-up.

## Where is the time now?

Profiling the log class in the same way:

```{r}
#| label: s3-profile
prof_file_log <- profile_expr(for (i in 1:15) fit(model_log))
prof_log <- summaryRprof(prof_file_log)
our_funs_log <- dQuote(c("em_step", "Q.tmix_log", "log_comp", "ldt_ls", "e_step.tmix_log",
                         "grad_Q.tmix", "plogis"), FALSE)
prof_log$by.total[rownames(prof_log$by.total) %in% our_funs_log, ]

calls_log <- count_calls(fit(model_log), c("e_step", "Q", "grad_Q"))
calls_log
```

- Each call to $Q$ is now much cheaper, but there are still **`r calls_log[["Q"]]` calls** per fit — the log scale changed the cost per call, not the number of calls.
- So $Q$ in the line search is still about `r round(prof_log$by.total['"Q.tmix_log"', "total.pct"])`% of the time, now spent in `log1p()` inside `ldt_ls()` instead of in `dt()`.

[The profile pointed at `dt()`, and the log scale removed it: $Q$ became `r round(med[["Q_density"]] / med[["Q_log"]])` times cheaper, the whole fit `r round(as.numeric(bm_classes$median[1] / bm_classes$median[2]))` times faster, and the answer is the same. The S3 design made the comparison fair: one GEM algorithm and one test suite for both implementations. The next hotspot is hotspot 1 — the number of $Q$ evaluations in the line search.]{.lesson}
