---
title: "EM for Mixtures of t-Distributions"
subtitle: "Computational Statistics — Assignment 3, topic B"
format:
  revealjs:
    theme: default
    css: styles.css
    slide-number: true
    smaller: true
    scrollable: true
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
    width: 1200
    height: 750
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    fig.width: 10
    fig.height: 4.6
    fig.align: center
    comment: "#>"
---

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(testthat)
source("R/load-v1.R")
theme_set(theme_minimal(base_size = 15))
options(digits = 5, width = 110)

# print the source of a function (with comments) in a code chunk
show_fun <- function(name) {
  paste0(name, " <- ", paste(as.character(attr(get(name), "srcref")), collapse = "\n"))
}
```

## Roadmap

1. **A simple implementation** — the formulas coded directly, on the density scale, with R's `dt()`
2. **Is it correct?** — a test against `optim()`
3. **Where is it slow?** — profiling with `Rprof` and `profvis`
4. **A faster implementation** — a log-scale version with a shorter first step in the line search, and an S3 design with two classes so the two implementations share the same GEM code
5. **How good is it?** — `bench::press` against R's own optimizers

# A simple implementation

## Two-component mixture of $t$-distributions

$$
f(x \mid \theta) = p\, f(x \mid \mu_1, \sigma_1^2, \nu_1) + (1 - p)\, f(x \mid \mu_2, \sigma_2^2, \nu_2),
\qquad \theta = (p, \mu_1, \mu_2, \sigma_1, \sigma_2)
$$

with $f(x \mid \mu, \sigma^2, \nu) = \frac{\Gamma((\nu+1)/2)}{\sqrt{\pi\nu\sigma^2}\,\Gamma(\nu/2)}\left(1 + \frac{(x-\mu)^2}{\nu\sigma^2}\right)^{-(\nu+1)/2}$ and $\nu_1, \nu_2$ fixed.

:::: {.columns}
::: {.column width="45%"}
Simulated data as in the assignment:

```{r}
#| label: sim-data
set.seed(42)
nu <- c(5, 10)
par_true <- c(p = 0.6, mu1 = 0, mu2 = 3,
              sigma1 = 1, sigma2 = 0.7)
x <- rtmix(400, par_true, nu)
par0 <- default_start(x)
round(par0, 3)
```

`default_start()`: $p = 0.5$, quartiles for $\mu$, $\text{sd}(x)/2$ for $\sigma$.
:::

::: {.column width="55%"}
```{r}
#| label: plot-data
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.3
grid <- seq(min(x), max(x), length.out = 400)
ggplot(data.frame(x), aes(x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "grey85", colour = "white") +
  geom_line(data = data.frame(x = grid, d = dtmix(grid, par_true, nu)), aes(x, d), linewidth = 1) +
  labs(y = "density", title = "N = 400, true density")
```
:::
::::

## Missing data: the component labels

**Idea:** if we knew which component each $x_i$ came from, estimation would be easy — just like the two coins in lecture 9.

Introduce the (unobserved) label $Z_i \in \{1, 2\}$:
$$
P(Z_i = 1) = p, \qquad X_i \mid Z_i = j \sim t(\mu_j, \sigma_j^2, \nu_j).
$$

Write $f_j(x) = f(x \mid \mu_j, \sigma_j^2, \nu_j)$. The joint density of one complete observation $(x_i, z_i)$ is
$$
f(x_i, z_i \mid \theta) = \big(p\, f_1(x_i)\big)^{\mathbf{1}(z_i = 1)} \big((1 - p)\, f_2(x_i)\big)^{\mathbf{1}(z_i = 2)}.
$$

. . .

**Check:** summing over $z_i \in \{1, 2\}$ gives $p f_1(x_i) + (1-p) f_2(x_i)$ — exactly the mixture density. So $(X, Z)$ is a valid complete-data model for our observed $X$.

**Complete-data log-likelihood** (what we could maximize if $Z$ were observed):
$$
\ell_c(\theta) = \sum_{i=1}^N \mathbf{1}(Z_i = 1)\big(\log p + \log f_1(x_i)\big) + \mathbf{1}(Z_i = 2)\big(\log(1-p) + \log f_2(x_i)\big)
$$

::: notes
The trick of EM is to choose missing data such that the complete-data problem is easy. Here the missing data are the labels. The check that summing out z gives the mixture density is what makes this a legal choice of complete data.
:::

## E-step: from $\ell_c$ to $Q$

$Q(\theta \mid \hat\theta) = E_{\hat\theta}\big(\ell_c(\theta) \mid X = x\big)$ — average $\ell_c$ over what the labels could be, given the data and the current guess $\hat\theta$.

**Step 1 — linearity.** Given $x$, the terms $\log p + \log f_1(x_i)$ are just numbers, so the expectation only hits the indicators:
$$
E_{\hat\theta}\big(\mathbf{1}(Z_i = 1) \mid X = x\big) = P_{\hat\theta}(Z_i = 1 \mid X_i = x_i) =: \gamma_i
$$

**Step 2 — Bayes' formula.** With $\hat f_j$ the density of component $j$ at the current guess $(\hat\mu_j, \hat\sigma_j)$:
$$
\gamma_i = \frac{P(Z_i = 1)\, f(x_i \mid Z_i = 1)}{f(x_i)} = \frac{\hat p\, \hat f_1(x_i)}{\hat p\, \hat f_1(x_i) + (1 - \hat p)\, \hat f_2(x_i)}
$$

**Result:**
$$
Q(\theta \mid \hat\theta) = \sum_{i=1}^N \gamma_i \big(\log p + \log f_1(x_i)\big) + (1 - \gamma_i)\big(\log(1-p) + \log f_2(x_i)\big)
$$

The $\gamma_i$ are computed from the **old** $\hat\theta$ and are fixed numbers ("soft labels"); $\theta$ in the logs is the **new** parameter we optimize over.

```{r}
#| label: plot-gamma
#| echo: false
#| fig-height: 2.6
par_hat <- gem(x, nu, eps = 1e-14)
gdf <- rbind(
  data.frame(x = grid, gamma = e_step(par0, grid, nu), theta = "at the start value"),
  data.frame(x = grid, gamma = e_step(par_hat, grid, nu), theta = "at the estimate")
)
ggplot(gdf, aes(x, gamma, colour = theta)) +
  geom_line(linewidth = 1) +
  geom_rug(data = data.frame(x), aes(x), inherit.aes = FALSE, alpha = 0.3) +
  labs(y = expression(gamma == P(Z == 1 ~ "|" ~ X == x)), colour = NULL)
```

## M-step: maximize $Q$ — what can be solved?

$Q$ splits into three parts that can be maximized separately:
$$
Q = \underbrace{\sum_i \gamma_i \log p + (1 - \gamma_i)\log(1-p)}_{\text{only } p}
  + \underbrace{\sum_i \gamma_i \log f_1(x_i)}_{\text{only } \mu_1, \sigma_1}
  + \underbrace{\sum_i (1 - \gamma_i) \log f_2(x_i)}_{\text{only } \mu_2, \sigma_2}
$$

**$p$:** $\ \partial_p Q = \frac{\sum_i \gamma_i}{p} - \frac{N - \sum_i \gamma_i}{1 - p} = 0 \;\Rightarrow\; p = \frac{1}{N}\sum_i \gamma_i$ — closed form.

. . .

**$(\mu_1, \sigma_1)$:** a *weighted* $t$ log-likelihood. With $r = x - \mu_1$ and $\log f_1(x) = c(\nu_1) - \log \sigma_1 - \frac{\nu_1 + 1}{2}\log\big(1 + \frac{r^2}{\nu_1 \sigma_1^2}\big)$, the chain rule gives
$$
\partial_{\mu_1} \log f_1(x) = \frac{w\, r}{\sigma_1^2},
\qquad
\partial_{\sigma_1} \log f_1(x) = -\frac{1}{\sigma_1} + \frac{w\, r^2}{\sigma_1^3},
\qquad w = \frac{\nu_1 + 1}{\nu_1 + r^2/\sigma_1^2},
$$
so
$$
\partial_{\mu_1} Q = \sum_i \gamma_i \frac{w_{i1} r_{i1}}{\sigma_1^2}, \qquad
\partial_{\sigma_1} Q = \sum_i \frac{\gamma_i}{\sigma_1}\left(\frac{w_{i1} r_{i1}^2}{\sigma_1^2} - 1\right) \qquad (\text{component 2: } 1 - \gamma_i).
$$

. . .

Setting $\partial_{\mu_1} Q = 0$ and $\partial_{\sigma_1} Q = 0$ gives
$$
\mu_1 = \frac{\sum_i \gamma_i w_{i1} x_i}{\sum_i \gamma_i w_{i1}}, \qquad
\sigma_1^2 = \frac{\sum_i \gamma_i w_{i1} r_{i1}^2}{\sum_i \gamma_i}
$$
— but $w_{i1}$ depends on $\mu_1$ and $\sigma_1$ themselves: fixed-point equations, not formulas.

[No closed-form M-step for $(\mu, \sigma)$. **Generalized EM:** we only need $Q(\theta_{n+1} \mid \theta_n) \ge Q(\theta_n \mid \theta_n)$, so: exact update of $p$ + one uphill (gradient) step for $(\mu_1, \mu_2, \sigma_1, \sigma_2)$.]{.problem}


## One gradient step — but how long?

First iteration from `par0`: E-step, update $p$, and then move a step of length $s$ along the gradient, $\theta + s\, g$. How much does $Q$ change?

```{r}
#| label: fixed-steps
#| echo: false
gam <- e_step(par0, x, nu)
par1 <- par0
par1[1] <- mean(gam)                       # exact M-step for p
g <- grad_Q(par1, gam, x, nu)
g[1] <- 0
change_in_Q <- function(s) Q(par1 + s * g, gam, x, nu) - Q(par1, gam, x, nu)
```

```{r}
#| label: plot-fixed-steps
#| echo: false
#| fig-height: 3.2
step_df <- data.frame(s = seq(0, 0.0085, length.out = 300))
step_df$change <- sapply(step_df$s, change_in_Q)
s_best <- optimize(change_in_Q, c(0, 0.0085), maximum = TRUE)$maximum
s_zero <- uniroot(change_in_Q, c(s_best, 0.0085))$root
ggplot(step_df, aes(s, change)) +
  annotate("rect", xmin = s_zero, xmax = Inf, ymin = -Inf, ymax = Inf, fill = "#d95f02", alpha = 0.12) +
  geom_hline(yintercept = 0, colour = "grey40") +
  geom_line(linewidth = 1.1, colour = "#2c7fb8") +
  geom_vline(xintercept = s_best, linetype = 2, colour = "grey40") +
  annotate("point", x = s_best, y = change_in_Q(s_best), size = 3) +
  annotate("text", x = s_best, y = change_in_Q(s_best) + 0.9, label = "best step", size = 4.5) +
  annotate("text", x = 0.0004, y = -2.2, label = "too short:\nbarely moves", size = 4.5, hjust = 0) +
  annotate("text", x = s_zero + 0.0002, y = 2.2, label = "too long:\nQ decreases", size = 4.5, hjust = 0, colour = "#d95f02") +
  labs(x = "step size s", y = "change in Q")
```

[The gradient only gives the *direction*. A long step jumps past the top and $Q$ **decreases** (then it is not a GEM step), and an even longer one ($s > `r signif(min(-par1[4:5] / g[4:5]), 2)`$) makes $\sigma$ negative; too short a step barely moves. The right length depends on the data: the gradient is a sum over $N$ points and scales like $N / \sigma^2$.]{.problem}

. . .

**Solution — backtracking:** start with a long step and halve it until $Q$ has increased enough (Armijo condition $Q(\theta + s g) \ge Q(\theta) + c\, s\, \|g\|^2$).

## Running it

:::: {.columns}
::: {.column width="45%"}
```{r}
#| label: run-gem
fit <- gem(x, nu)
fit
loglik(fit, x, nu)
```
:::
::: {.column width="55%"}
```{r}
#| label: plot-fit
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.3
p <- fit
df_fit <- data.frame(
  x = rep(grid, 4),
  density = c(dtmix(grid, p, nu), p[1] * dt_ls(grid, p[2], p[4], nu[1]), (1 - p[1]) * dt_ls(grid, p[3], p[5], nu[2]),
              dtmix(grid, par_true, nu)),
  curve = rep(c("mixture", "component 1", "component 2", "true density"), each = length(grid))
)
ggplot(data.frame(x), aes(x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "grey85", colour = "white") +
  geom_line(data = df_fit, aes(x, density, colour = curve, linetype = curve), linewidth = 0.9) +
  scale_colour_manual(values = c("#F8766D", "#00BA38", "#619CFF", "black")) +
  scale_linetype_manual(values = c(2, 2, 1, 3)) +
  labs(colour = NULL, linetype = NULL, y = "density", title = "Fitted mixture and true density")
```
:::
::::

# Is it correct?

## A test against `optim()`

`optim()` is R's general-purpose optimizer. It knows nothing about EM: it just maximizes the observed log-likelihood $\ell(\theta)$ directly. If my GEM is implemented correctly, the two must find the same maximum.

Both are started at the same point (`default_start()`):

```{r}
#| label: test-optim
#| echo: false
nll <- function(eta) -loglik(c(plogis(eta[1]), eta[2:3], exp(eta[4:5])), x, nu)
eta <- optim(c(qlogis(par0[1]), par0[2:3], log(par0[4:5])), nll,
             method = "BFGS", control = list(reltol = 1e-14, maxit = 1000))$par
par_optim <- c(plogis(eta[1]), eta[2:3], exp(eta[4:5]))

optim_tab <- rbind(
  GEM = c(par_hat, loglik(par_hat, x, nu)),
  `optim() (BFGS)` = c(par_optim, loglik(par_optim, x, nu))
)
optim_fmt <- rbind(
  formatC(optim_tab, format = "f", digits = 4),
  Difference = formatC(abs(optim_tab[1, ] - optim_tab[2, ]), format = "e", digits = 1)
)
knitr::kable(
  optim_fmt,
  col.names = c("$p$", "$\\mu_1$", "$\\mu_2$", "$\\sigma_1$", "$\\sigma_2$", "$\\ell(\\hat\\theta)$"),
  align = "r"
)
```

Two different algorithms agree on the estimate and on the log-likelihood.

# Where is it slow?

## Why? Number of calls × cost per call

```{r}
#| label: prof-setup
#| include: false
set.seed(7)
x_big <- rtmix(5000, par_true, nu)
fit_big <- gem(x_big, nu)
n_iter <- attr(fit_big, "iter")
bm_fit <- bench::mark(gem(x_big, nu), iterations = 5)

prof_file <- profile_expr(for (i in 1:5) gem(x_big, nu))
prof <- summaryRprof(prof_file)
```

The profile says *where* the time goes. To see *why*, count how often each building block is called in one fit ($N = 5000$, `r n_iter` iterations) and time a single call with `bench::mark`:

```{r}
#| label: prof-calls
#| echo: false
calls <- count_calls(gem(x_big, nu), c("e_step", "Q", "grad_Q", "dt_ls"))

gam_big <- e_step(fit_big, x_big, nu)
cost <- bench::mark(
  e_step = e_step(fit_big, x_big, nu),
  Q      = Q(fit_big, gam_big, x_big, nu),
  grad_Q = grad_Q(fit_big, gam_big, x_big, nu),
  dt_ls  = dt_ls(x_big, 0, 1, 5),
  check = FALSE
)
per_call <- setNames(as.numeric(cost$median), as.character(cost$expression))[names(calls)]
total_ms <- 1000 * calls * per_call
calls_tab <- data.frame(
  fun   = paste0("`", names(calls), "()`"),
  calls = formatC(calls, format = "d"),
  per_iter = formatC(calls / n_iter, format = "f", digits = 1),
  ms_call  = formatC(1000 * per_call, format = "f", digits = 2),
  total    = formatC(total_ms, format = "f", digits = 0),
  share    = paste0(round(100 * total_ms / (1000 * as.numeric(bm_fit$median))), "%")
)
calls_tab <- calls_tab[c(2, 1, 3, 4), ]
calls_tab[1, ] <- paste0("**", calls_tab[1, ], "**")
calls_tab$fun[4] <- paste0(calls_tab$fun[4], " *(inside `Q()` and `e_step()`)*")
knitr::kable(
  calls_tab, row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
  col.names = c("Function", "Calls per fit", "Calls per iteration", "Time per call (ms)",
                "Total time (ms)", "Share of the fit")
)
```

- One call to `Q()` costs about the same as one E-step — but `Q()` is called **`r round(calls[["Q"]] / n_iter, 1)` times per iteration**, the E-step once.
- `dt_ls()` is not an extra cost: it is the density inside `Q()` and `e_step()` (2 calls in each), so its row overlaps with theirs.
- The gradient is almost free.

## Conclusion: three hotspots

1. **The line search.** $Q$ is evaluated `r round(calls[["Q"]] / n_iter, 1)` times per iteration (`r calls[["Q"]]` calls in `r n_iter` iterations), because the backtracking restarts from step size 1 in *every* iteration and needs about `r round(calls[["Q"]] / n_iter - 2)` halvings before the step is short enough ($\approx 2^{-`r round(calls[["Q"]] / n_iter - 2)`}$).
2. **`dt()`.** Most of the time is spent inside R's `dt()`. Every $Q$ and every E-step evaluates the $t$-density twice for all $N$ points: `r calls[["dt_ls"]]` density evaluations of length `r length(x_big)` per fit.
3. **Repeated work.** The E-step computes $p f_1(x_i)$ and $(1-p) f_2(x_i)$; right after, $Q$ (at the same parameters, as `Q_old`) computes the same densities again, and `grad_Q()` recomputes the residuals.

[The profile is dominated by one line (the $Q$ call in the line search) and one primitive (`dt()`). Counting calls and timing each building block explains *why*: total time ≈ number of calls × cost per call.]{.lesson}

::: notes
Transition: first a look at the step sizes behind hotspot 1, which gives a cheap fix (start the line search at 2^-5). Hotspot 2 (dt) is then attacked with a log-scale implementation. The new profile at the end shows what is left of hotspot 1.
:::

## How long are the steps actually?

```{r}
#| label: step-sizes
#| echo: false
# Replay the GEM iterations and recover the accepted step size s in each one:
# par_new = par + s * g, so s = <par_new - par, g> / ||g||^2.
step_sizes <- function(x, nu, par = default_start(x), eps = 1e-10, maxit = 1000) {
  gem_step <- create_gem_step(x, nu)
  out <- list()
  for (i in seq_len(maxit)) {
    gam <- e_step(par, x, nu)
    par1 <- par
    par1[1] <- mean(gam)
    g <- grad_Q(par1, gam, x, nu)
    g[1] <- 0
    par_new <- gem_step(par)
    out[[i]] <- data.frame(iter = i, step = sum((par_new - par1) * g) / sum(g^2), grad = sqrt(sum(g^2)))
    done <- sum((par_new - par)^2) <= eps * (sum(par_new^2) + eps)
    par <- par_new
    if (done) break
  }
  do.call(rbind, out)
}
steps <- rbind(
  cbind(N = "N = 400", step_sizes(x, nu)),
  cbind(N = "N = 5000", step_sizes(x_big, nu))
)
steps_long <- rbind(
  data.frame(steps[c("N", "iter")], what = "step size s", value = steps$step),
  data.frame(steps[c("N", "iter")], what = "length of the move, s * ||g||", value = steps$step * steps$grad)
)
steps_long$what <- factor(steps_long$what, levels = unique(steps_long$what))
```

```{r}
#| label: plot-step-sizes
#| echo: false
#| fig-height: 3.6
ggplot(steps_long, aes(iter, value, colour = N)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 2) +
  facet_wrap(~what, scales = "free_y") +
  scale_y_continuous(trans = "log2", labels = function(v) parse(text = paste0("2^", round(log2(v))))) +
  scale_colour_manual(values = c("#2c7fb8", "#d95f02")) +
  labs(x = "iteration", y = NULL, colour = NULL) +
  theme(legend.position = "top")
```

- The accepted step size is almost **constant**: $2^{-8} \approx `r signif(2^-8, 2)`$ for $N = 400$ (one iteration needs $2^{-9}$) and $2^{-11}$ or $2^{-12}$ for $N = 5000$ — it scales like $1/N$, because the gradient is a sum over $N$ points.
- The *move* $s\,\|g\|$ shrinks only because the gradient does. Starting the backtracking at $s = 1$ in every iteration is wasted work.

[The first halvings are always rejected. Version 2 therefore starts the backtracking at $s = 2^{-5}$: still well above every accepted step, so the line search adapts to $N$ as before, but the five longest steps are never tried.]{.lesson}

# A faster implementation: the log scale

```{r}
#| label: load-s3
#| include: false
fit_v1 <- fit # the version 1 estimate, before fit() becomes a generic
# Version 2 of the code: the building blocks become S3 generics (replaces the plain functions)
source("R/load.R")
```

## From the profile to a plan

**What the profile showed:** `Q()` takes `r round(prof$by.total['"Q"', "total.pct"])`% of the time, and `dt()` alone `r round(prof$by.self['"dt"', "self.pct"])`%.

**Observation:** $Q$ only needs **log**-densities,
$$
Q(\theta \mid \hat\theta) = \sum_i \gamma_i \big(\log p + \log f_1(x_i)\big) + (1 - \gamma_i)\big(\log(1-p) + \log f_2(x_i)\big),
$$
but version 1 computes the density with `dt()` and then takes `log()` of it — a detour through the density scale.

. . .

**Idea:** write the log-density out and compute it directly,
$$
\log f(x) = \underbrace{\log\Gamma\big(\tfrac{\nu+1}{2}\big) - \log\Gamma\big(\tfrac{\nu}{2}\big) - \tfrac12\log(\pi\nu)}_{c(\nu)\text{: a constant, computed once}} - \log\sigma - \tfrac{\nu+1}{2}\log\Big(1 + \tfrac{(x-\mu)^2}{\nu\sigma^2}\Big).
$$
With $a_i = \log p + \log f_1(x_i)$ and $b_i = \log(1-p) + \log f_2(x_i)$:

| Building block | Version 1 (density scale) | Log scale |
|---|---|---|
| $Q$ | `log(p * dt(...))`, twice per point | $\sum_i \gamma_i a_i + (1-\gamma_i) b_i$ — no `dt()`, no `log()` |
| E-step | $\gamma_i = \frac{p f_1}{p f_1 + (1-p) f_2}$ | $\gamma_i = \frac{1}{1 + e^{b_i - a_i}}$ = `plogis(a - b)` |
| log-likelihood | `sum(log(dtmix(...)))` | $\log(e^{a_i} + e^{b_i})$ with the log-sum-exp trick |

. . .

**And hotspot 1:** the line search now starts at $s = 2^{-5}$ instead of $1$ (`step0 = 2^-5` in `create_gem_step()`).

**To compare the two fairly** they must run through the *same* GEM code and the *same* tests. So the code is restructured with S3: one class per implementation, and everything they share in a common parent class.

## The S3 design: two classes with a common parent

:::: {.columns}
::: {.column width="48%"}
```{=html}
<svg viewBox="0 0 520 400" style="width: 100%; max-width: 540px;" xmlns="http://www.w3.org/2000/svg" font-family="monospace" font-size="15" role="img" aria-label="Class diagram: tmix is the parent of tmix_density and tmix_log">
  <g stroke="#2c7fb8" stroke-width="1.5" fill="white">
    <rect x="150" y="8" width="220" height="150"/>
    <rect x="150" y="8" width="220" height="32" fill="#dcebf7"/>
    <line x1="150" y1="92" x2="370" y2="92"/>
    <rect x="8" y="230" width="230" height="162"/>
    <rect x="8" y="230" width="230" height="32" fill="#dcebf7"/>
    <line x1="8" y1="294" x2="238" y2="294"/>
    <rect x="282" y="230" width="230" height="162"/>
    <rect x="282" y="230" width="230" height="32" fill="#dcebf7"/>
    <line x1="282" y1="294" x2="512" y2="294"/>
    <polyline points="123,230 123,200 215,200 215,172" fill="none"/>
    <polygon points="215,158 207,172 223,172"/>
    <polyline points="397,230 397,200 305,200 305,172" fill="none"/>
    <polygon points="305,158 297,172 313,172"/>
  </g>
  <g fill="#222">
    <text x="260" y="30" text-anchor="middle" font-weight="bold">tmix</text>
    <text x="162" y="62">x   (data)</text>
    <text x="162" y="82">nu  (shape parameters)</text>
    <text x="162" y="114">grad_Q()</text>
    <text x="162" y="132">fit()</text>
    <text x="162" y="150">print()</text>
    <text x="123" y="252" text-anchor="middle" font-weight="bold">tmix_density</text>
    <text x="20" y="282" fill="#777">(no extra fields)</text>
    <text x="20" y="316">loglik()</text>
    <text x="20" y="334">e_step()</text>
    <text x="20" y="352">Q()</text>
    <text x="20" y="380" fill="#2c7fb8">density scale: dt()</text>
    <text x="397" y="252" text-anchor="middle" font-weight="bold">tmix_log</text>
    <text x="294" y="282">log_const  (= c(nu))</text>
    <text x="294" y="316">loglik()</text>
    <text x="294" y="334">e_step()</text>
    <text x="294" y="352">Q()</text>
    <text x="294" y="380" fill="#2c7fb8">log scale: log1p()</text>
  </g>
</svg>
```
:::
::: {.column width="52%"}
**`tmix`** (parent) holds what is the same for both:

- the data `x` and the shape parameters `nu`
- `grad_Q()`: the gradient is the same formula
- `fit()`: the GEM algorithm
- `print()`

**`tmix_density`**: version 1 — `loglik()`, `e_step()` and `Q()` with `dt()`.

**`tmix_log`**: the same three methods on log scale, plus the constant $c(\nu)$ stored in the object.

Only what *differs* between the implementations is written twice.
:::
::::

## Generics, methods and dispatch

A **generic** only dispatches; a **method** is an ordinary function named `generic.class`:

| Generic | `tmix_density` | `tmix_log` |
|---|---|---|
| `loglik()` | `loglik.tmix_density` | `loglik.tmix_log` |
| `e_step()` | `e_step.tmix_density` | `e_step.tmix_log` |
| `Q()` | `Q.tmix_density` | `Q.tmix_log` |
| `grad_Q()` | `grad_Q.tmix` (inherited) | `grad_Q.tmix` (inherited) |
| `fit()` | `fit.tmix` (inherited) | `fit.tmix` (inherited) | | `print()` | `print.tmix` (inherited) | `print.tmix` (inherited) |


## One GEM algorithm for both classes

The GEM code only talks to the model through the generics (`e_step(model, par)`, `Q(model, par, gam)`, `grad_Q(model, par, gam)`), so the same `fit()` works for both classes.

Make the two models and fit them:

```{r}
#| label: s3-fit
model_d <- tmix_density(x, nu)   # density scale, dt()
model_l <- tmix_log(x, nu)       # log scale

rbind(density = fit(model_d),
      log     = fit(model_l))
```

Same algorithm, same estimate — only the way the densities are computed differs. Both classes use the new first step $2^{-5}$; the halving passes through the same step sizes as before, so the estimate is also the same as in version 1 (`r formatC(max(abs(fit(model_l) - fit_v1)), format = "e", digits = 0)` apart).

## Do the two classes agree?

The density class is the tested version 1, so the log class is tested against it:

```{r}
#| label: s3-agree
par_test <- c(0.4, -0.5, 2.5, 1.3, 0.9)
gam_test <- e_step(model_d, par0)

test_that("the log class computes the same as the density class", {
  expect_equal(loglik(model_l, par_test), loglik(model_d, par_test))
  expect_equal(e_step(model_l, par_test), e_step(model_d, par_test))
  expect_equal(Q(model_l, par_test, gam_test), Q(model_d, par_test, gam_test))
  expect_equal(fit(model_l), fit(model_d), tolerance = 1e-8)
})
```

A bonus of the log scale: it survives an extreme outlier where the density scale underflows to $0/0$:

```{r}
#| label: s3-outlier
x_out <- c(x, -1e60)
c(density = tail(e_step(tmix_density(x_out, nu), par_true), 1),
  log     = tail(e_step(tmix_log(x_out, nu), par_true), 1))
```

## Is it faster?

The two classes timed with `bench::mark` on the data from the profiling ($N = 5000$). `bench::mark` also checks that both give the same estimate.

```{r}
#| label: s3-bench
#| echo: false
model_density <- tmix_density(x_big, nu)
model_log <- tmix_log(x_big, nu)
same_fit <- function(a, b) isTRUE(all.equal(as.vector(a), as.vector(b), tolerance = 1e-8))
bm_classes <- bench::mark(density = fit(model_density), log = fit(model_log),
                          check = same_fit, iterations = 10)

par_b <- fit(model_log)
gam_b <- e_step(model_log, par_b)
blocks <- bench::mark(
  Q_density = Q(model_density, par_b, gam_b),           Q_log = Q(model_log, par_b, gam_b),
  e_step_density = e_step(model_density, par_b),        e_step_log = e_step(model_log, par_b),
  loglik_density = loglik(model_density, par_b),        loglik_log = loglik(model_log, par_b),
  check = FALSE
)
med <- setNames(1e6 * as.numeric(blocks$median), as.character(blocks$expression))

ms_density <- c(1000 * as.numeric(bm_classes$median[1]), med[c("Q_density", "e_step_density", "loglik_density")] / 1000)
ms_log     <- c(1000 * as.numeric(bm_classes$median[2]), med[c("Q_log", "e_step_log", "loglik_log")] / 1000)
fmt_ms <- function(v) ifelse(v >= 10, formatC(v, format = "f", digits = 0), formatC(v, format = "f", digits = 2))
faster_tab <- data.frame(
  what    = c("Whole fit", "One call to `Q()`", "One call to `e_step()`", "One call to `loglik()`"),
  density = fmt_ms(ms_density),
  log     = fmt_ms(ms_log),
  speed   = paste0(formatC(ms_density / ms_log, format = "f", digits = 1), " ×")
)
faster_tab[1, ] <- paste0("**", faster_tab[1, ], "**")
knitr::kable(
  faster_tab, row.names = FALSE, align = c("l", "r", "r", "r"),
  col.names = c("", "`tmix_density` (ms)", "`tmix_log` (ms)", "Speed-up")
)
```

- The whole fit is **`r round(as.numeric(bm_classes$median[1] / bm_classes$median[2]), 1)` times faster** on log scale, with the same estimate and the same number of iterations. Both classes start the line search at $2^{-5}$, so this is the effect of the log scale alone.
- `Q()` got the largest speed-up — and `Q()` was what dominated the profile.

## Where is the time now?

Profiling the log class in the same way as before, and counting the calls again. Version 1 is the density scale with first step $1$; version 2 is the log scale with first step $2^{-5}$:

```{r}
#| label: s3-profile
#| echo: false
prof_file_log <- profile_expr(for (i in 1:15) fit(model_log))
prof_log <- summaryRprof(prof_file_log)
calls_log <- count_calls(fit(model_log), c("e_step", "Q", "grad_Q"))

share <- function(p, f) { v <- p$by.total[dQuote(f, FALSE), "total.pct"]; if (is.na(v)) 0 else v }
now_tab <- data.frame(
  fun = c("`Q()`", "`e_step()`", "`grad_Q()`"),
  calls_before = formatC(calls[c("Q", "e_step", "grad_Q")], format = "d"),
  calls_now    = formatC(calls_log[c("Q", "e_step", "grad_Q")], format = "d"),
  share_before = paste0(round(c(share(prof, "Q"), share(prof, "e_step"), share(prof, "grad_Q"))), "%"),
  share_now    = paste0(round(c(share(prof_log, "Q.tmix_log"), share(prof_log, "e_step.tmix_log"),
                                share(prof_log, "grad_Q.tmix"))), "%")
)
now_tab[1, ] <- paste0("**", now_tab[1, ], "**")
knitr::kable(
  now_tab, row.names = FALSE, align = c("l", "r", "r", "r", "r"),
  col.names = c("Function", "Calls per fit, version 1", "Calls per fit, version 2",
                "Share of time, version 1", "Share of time, version 2")
)
```

- **Fewer calls:** starting at $2^{-5}$ saves `r round((calls[["Q"]] - calls_log[["Q"]]) / n_iter)` evaluations of $Q$ per iteration — `r round(calls_log[["Q"]] / n_iter, 1)` instead of `r round(calls[["Q"]] / n_iter, 1)`, with the same `r n_iter` iterations. (Not all five skipped steps cost a call before: the longest ones made $\sigma$ negative and were rejected without evaluating $Q$.) **Cheaper calls:** the log scale.
- $Q$ in the line search is still about `r round(prof_log$by.total['"Q.tmix_log"', "total.pct"])`% of the time, now spent in `log1p()` inside `ldt_ls()` instead of in `dt()`: the step is still halved about `r round(calls_log[["Q"]] / n_iter - 2)` times per iteration, from $2^{-5}$ down to $2^{-11}$ or $2^{-12}$.

[The profile pointed at `dt()`, and the log scale removed it: $Q$ became `r round(med[["Q_density"]] / med[["Q_log"]])` times cheaper, the whole fit `r round(as.numeric(bm_classes$median[1] / bm_classes$median[2]))` times faster, and the answer is the same. The S3 design made the comparison fair: one GEM algorithm for both implementations. Starting the line search at $2^{-5}$ took `r round(100 * (1 - calls_log[["Q"]] / calls[["Q"]]))`% of the $Q$ evaluations away, but hotspot 1 is not gone: the line search still restarts in every iteration.]{.lesson}

# How good is it? Benchmark against R's optimizers

## The competitors

R has several general-purpose optimizers. They do not use the EM idea: they maximize the observed log-likelihood $\ell(\theta)$ directly.

| Method | Function | Type |
|---|---|---|
| GEM | `fit()` (ours) | E-step + one gradient step on $Q$ with backtracking |
| BFGS | `optim(method = "BFGS")` | quasi-Newton |
| L-BFGS-B | `optim(method = "L-BFGS-B")` | quasi-Newton, limited memory |
| CG | `optim(method = "CG")` | conjugate gradient |
| Nelder-Mead | `optim(method = "Nelder-Mead")` | simplex, uses no gradient |
| nlminb | `nlminb()` | quasi-Newton (PORT library) |

## Do they all find the same maximum?

On the data from the profiling ($N = 5000$), all started at `default_start()`:

```{r}
#| label: methods-accuracy
#| echo: false
methods <- list(
  GEM           = function(m) fit(m),
  BFGS          = function(m) fit_optim(m, "BFGS"),
  `L-BFGS-B`    = function(m) fit_optim(m, "L-BFGS-B"),
  CG            = function(m) fit_optim(m, "CG"),
  `Nelder-Mead` = function(m) fit_optim(m, "Nelder-Mead"),
  nlminb        = function(m) fit_nlminb(m)
)
fits <- lapply(methods, function(f) f(model_log))
ll <- sapply(fits, function(p) loglik(model_log, p))
max_tab <- data.frame(
  method = names(fits),
  formatC(t(sapply(fits, as.vector)), format = "f", digits = 4),
  evals  = formatC(sapply(fits, attr, "iter"), format = "d"),
  gap    = ifelse(max(ll) - ll == 0, "0", formatC(max(ll) - ll, format = "e", digits = 1))
)
max_tab$method[1] <- "**GEM (ours)**"
knitr::kable(
  max_tab, row.names = FALSE, align = c("l", rep("r", 7)),
  col.names = c("Method", "$p$", "$\\mu_1$", "$\\mu_2$", "$\\sigma_1$", "$\\sigma_2$",
                "Evaluations", "Distance to best $\\ell$")
)
```

Evaluations: iterations for the GEM, function evaluations for the others. The last column is the distance to the best log-likelihood found.

## `bench::press`: time as a function of $N$

```{r}
#| label: press-methods
#| echo: false
press <- bench::press(N = c(500, 2000, 8000, 32000), {
  set.seed(1)
  model <- tmix_log(rtmix(N, par_true, nu), nu)
  bench::mark(
    GEM           = fit(model),
    BFGS          = fit_optim(model, "BFGS"),
    `L-BFGS-B`    = fit_optim(model, "L-BFGS-B"),
    CG            = fit_optim(model, "CG"),
    `Nelder-Mead` = fit_optim(model, "Nelder-Mead"),
    nlminb        = fit_nlminb(model),
    check = FALSE, min_iterations = 3
  )
})
```

```{r}
#| label: plot-press-methods
#| echo: false
#| fig-height: 3.6
press_df <- data.frame(method = as.character(press$expression), N = press$N,
                       ms = 1000 * as.numeric(press$median))
ggplot(press_df, aes(N, ms, colour = method)) +
  geom_line(linewidth = 0.9) + geom_point(size = 2) +
  scale_x_log10(breaks = unique(press_df$N)) + scale_y_log10() +
  labs(y = "median time (ms)", colour = NULL)
```

## The same numbers as a table

```{r}
#| label: press-table
#| echo: false
tab <- reshape(press_df, idvar = "method", timevar = "N", direction = "wide")
names(tab) <- sub("ms\\.", "N = ", names(tab))
tab[order(tab[[ncol(tab)]]), ] |> transform(method = method) |>
  (\(d) { d[-1] <- round(d[-1], 1); rownames(d) <- NULL; d })()
```

Median time in ms, sorted by the time for the largest $N$.

- **The methods scale linearly in $N$** (parallel lines): they spend their time on passes over the data, and the number of passes hardly depends on $N$. The exception is CG, which is erratic: the number of evaluations it needs varies a lot from data set to data set.
- **The quasi-Newton methods are fastest.** They use curvature information and need few evaluations of $\ell$.
- **The GEM is clearly slower than the quasi-Newton methods.** It converges linearly, and each iteration costs about `r round(calls_log[["Q"]] / n_iter)` evaluations of $Q$ in the line search, even with the shorter first step — hotspot 1 from the profile.
- **Nelder-Mead is slowest of the stable methods**: it uses no gradient and needs hundreds of evaluations.

# Robustness: $t$ mixture vs Gaussian mixture

## A Gaussian mixture for comparison

The assignment suggests comparing the location and scale estimates with those from a two-component **Gaussian** mixture, in particular with outliers in the sample. The Gaussian mixture has a closed-form M-step, so it is an ordinary EM algorithm, run with the same driver `em()`:

```{r}
#| label: gauss-step
#| echo: false
#| results: asis
cat("```r\n", show_fun("create_em_gauss_step"), "\n```\n", sep = "")
```

Contaminate the sample from before ($N = 400$) with 10 outliers $\pm U(8, 20)$, i.e. 2.4% of the data, and fit both models from the same starting value (the true parameters):

```{r}
#| label: gauss-one
#| echo: false
set.seed(1)
x_cont <- c(x, runif(10, 8, 20) * sample(c(-1, 1), 10, replace = TRUE))
one_tab <- rbind(
  `True parameters`              = par_true,
  `$t$ mixture, clean`           = fit(tmix_log(x, nu), par0 = par_true),
  `Gaussian mixture, clean`      = fit_gauss(x, par0 = par_true),
  `$t$ mixture, 10 outliers`     = fit(tmix_log(x_cont, nu), par0 = par_true),
  `Gaussian mixture, 10 outliers` = fit_gauss(x_cont, par0 = par_true)
)
knitr::kable(
  one_tab, digits = 2, align = "r",
  col.names = c("$p$", "$\\mu_1$", "$\\mu_2$", "$\\sigma_1$", "$\\sigma_2$")
)
```

## The two fits on the contaminated sample

```{r}
#| label: plot-gauss-fit
#| echo: false
#| fig-height: 4.4
dgmix <- function(x, par) par[1] * dnorm(x, par[2], par[4]) + (1 - par[1]) * dnorm(x, par[3], par[5])
fit_t_cont <- one_tab[4, ]
fit_g_cont <- one_tab[5, ]
grid_cont <- seq(min(x_cont) - 1, max(x_cont) + 1, length.out = 1000)
df_cont <- data.frame(
  x = rep(grid_cont, 5),
  density = c(
    dgmix(grid_cont, fit_g_cont),
    fit_g_cont[1] * dnorm(grid_cont, fit_g_cont[2], fit_g_cont[4]),
    (1 - fit_g_cont[1]) * dnorm(grid_cont, fit_g_cont[3], fit_g_cont[5]),
    dtmix(grid_cont, fit_t_cont, nu),
    dtmix(grid_cont, par_true, nu)
  ),
  curve = factor(rep(c("Gaussian mixture", "Gaussian component 1", "Gaussian component 2",
                       "t mixture", "true density"), each = length(grid_cont)),
                 levels = c("Gaussian mixture", "Gaussian component 1", "Gaussian component 2",
                            "t mixture", "true density"))
)
ggplot(data.frame(x = x_cont), aes(x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 80, fill = "grey85", colour = "white") +
  geom_rug(data = data.frame(x = tail(x_cont, 10)), colour = "#d95f02", linewidth = 1,
           length = unit(0.06, "npc")) +
  geom_line(data = df_cont, aes(x, density, colour = curve, linetype = curve), linewidth = 0.9) +
  scale_colour_manual(values = c("#d95f02", "#d95f02", "#d95f02", "#2c7fb8", "black")) +
  scale_linetype_manual(values = c(1, 2, 3, 1, 2)) +
  labs(colour = NULL, linetype = NULL, y = "density",
       title = "N = 400 and 10 outliers (orange marks)")
```

- The Gaussian mixture has no heavy tails, so the only way it can give the outliers a positive likelihood is to **widen component 1** ($\sigma_1 = `r formatC(fit_g_cont[4], format = "f", digits = 2)`$). That component now lies under the whole sample, its mean is pulled to `r formatC(fit_g_cont[2], format = "f", digits = 2)`, and it takes weight from component 2 ($p = `r formatC(fit_g_cont[1], format = "f", digits = 2)`$).
- The result is that the left mode has disappeared: the fitted Gaussian density is far too low around 0 and too high on the flanks, where there are almost no data.
- The $t$ mixture keeps both modes and stays much closer to the true density: its tails already allow for a few far-away observations, so it only widens component 1 a little.

## Estimates as a function of the number of outliers

20 simulated data sets ($N = 400$). To each, add $k$ outliers $\pm U(8, 20)$ and fit both mixtures, started at the true parameters so that only the *models* differ:

```{r}
#| label: gauss-outliers
#| echo: false
set.seed(3)
k_out <- c(0, 2, 5, 10, 20, 40)
out_res <- do.call(rbind, lapply(1:20, function(rep) {
  x0 <- rtmix(400, par_true, nu)
  out <- runif(max(k_out), 8, 20) * sample(c(-1, 1), max(k_out), replace = TRUE)
  do.call(rbind, lapply(k_out, function(k) {
    xc <- c(x0, out[seq_len(k)])
    rbind(
      data.frame(k, model = "t mixture", t(fit(tmix_log(xc, nu), par0 = par_true))),
      data.frame(k, model = "Gaussian mixture", t(fit_gauss(xc, par0 = par_true)))
    )
  }))
}))
out_med <- aggregate(cbind(mu1, mu2, sigma1, sigma2) ~ k + model, out_res, median)
out_long <- reshape(out_med, direction = "long", varying = par_names[-1], v.names = "estimate",
                    timevar = "parameter", times = par_names[-1])
med_at <- function(model, k, par) out_med[out_med$model == model & out_med$k == k, par]
```

```{r}
#| label: plot-gauss-outliers
#| echo: false
#| fig-height: 3.4
ggplot(out_long, aes(k, estimate, colour = model)) +
  geom_hline(data = data.frame(parameter = par_names[-1], value = par_true[-1]),
             aes(yintercept = value), linetype = 2) +
  geom_line(linewidth = 0.9) + geom_point(size = 2) +
  facet_wrap(~parameter, nrow = 1, scales = "free_y") +
  labs(x = "number of outliers", y = "median estimate", colour = NULL) +
  theme(legend.position = "bottom")
```

Median estimate over the 20 data sets; the dashed lines are the true values.

- **Without outliers** the two models agree on the locations, but the Gaussian mixture overestimates $\sigma_1$ (`r formatC(med_at("Gaussian mixture", 0, "sigma1"), format = "f", digits = 2)` against the true 1): it has to explain the heavy tails of the $t_5$ component with a larger variance.
- **Two outliers (0.5%) are enough** to move the Gaussian $\sigma_1$ to `r formatC(med_at("Gaussian mixture", 2, "sigma1"), format = "f", digits = 2)` and $\mu_1$ to `r formatC(med_at("Gaussian mixture", 2, "mu1"), format = "f", digits = 2)`; the $t$ mixture gives `r formatC(med_at("t mixture", 2, "sigma1"), format = "f", digits = 2)` and `r formatC(med_at("t mixture", 2, "mu1"), format = "f", digits = 2)`.
- **With 20 or more outliers the Gaussian mixture breaks down**: one component is used for the outliers alone ($\sigma_1 \approx `r round(med_at("Gaussian mixture", 20, "sigma1"))`$) and the other has to cover both real components.
- The $t$ mixture degrades gradually instead. With fixed $\nu$ it is not immune: $\sigma_1$ still grows with the number of outliers.

[The heavy tails of the $t$-distribution make the estimates robust: in the gradient an observation far from $\mu$ gets the weight $w = (\nu + 1) / (\nu + r^2 / \sigma^2)$, which goes to 0, where the Gaussian mixture gives every observation full weight.]{.lesson}
