---
title: "EM for Mixtures of t-Distributions"
subtitle: "Computational Statistics — Assignment 3, topic B"
format:
  revealjs:
    theme: default
    css: styles.css
    slide-number: true
    smaller: true
    scrollable: true
    code-overflow: wrap
    highlight-style: github
    embed-resources: true
    width: 1200
    height: 750
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    fig.width: 10
    fig.height: 4.6
    fig.align: center
    comment: "#>"
---

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(patchwork)
library(testthat)
source("R/load.R")
theme_set(theme_minimal(base_size = 15))
options(digits = 5, width = 110)

# print the source of a function (with comments) in a code chunk
show_fun <- function(name) {
  paste0(name, " <- ", paste(as.character(attr(get(name), "srcref")), collapse = "\n"))
}
method_cols <- c(
  em_aug = "#1b9e77", em_exact = "#7570b3", gem = "#d95f02",
  gem_cold = "#e6ab02", gd = "#e7298a", bfgs = "#666666"
)
```

## Roadmap

1. **The model** and deriving the EM algorithm
2. **Version 1 → version 2**: a generalized EM that actually *is* one
3. **Testing**: how do I know it is correct?
4. **Alternatives**: augmented EM, exact EM, gradient ascent, BFGS
5. **Code structure**: one driver, many step functions, an S3 class
6. **Convergence criteria and benchmarking** (`bench::mark`, `bench::press`)
7. **Profiling** (`Rprof`/`profvis`) and optimization (R → Rcpp)
8. **Theory check**: Fisher information and convergence rates
9. **Robustness**: starting values, many data sets, outliers

Throughout: [boxes like this one]{.lesson} mark what I learned along the way.

::: notes
The presentation follows the order in which I actually worked on the problem.
Each "What I learned" box is a point where a test, a benchmark or a profile
changed my implementation.
:::

# The model

## Two-component mixture of $t$-distributions

$$
f(x \mid \theta) = p\, f(x \mid \mu_1, \sigma_1^2, \nu_1) + (1 - p)\, f(x \mid \mu_2, \sigma_2^2, \nu_2),
\qquad \theta = (p, \mu_1, \mu_2, \sigma_1, \sigma_2)
$$

with $f(x \mid \mu, \sigma^2, \nu) = \frac{\Gamma((\nu+1)/2)}{\sqrt{\pi\nu\sigma^2}\,\Gamma(\nu/2)}\left(1 + \frac{(x-\mu)^2}{\nu\sigma^2}\right)^{-(\nu+1)/2}$ and $\nu_1, \nu_2$ fixed.

:::: {.columns}
::: {.column width="45%"}
Simulated data as in the assignment:

```{r}
#| label: sim-data
set.seed(42)
nu <- c(5, 10)
par_true <- c(p = 0.6, mu1 = 0, mu2 = 3,
              sigma1 = 1, sigma2 = 0.7)
x <- rtmix(400, par_true, nu)
par0 <- default_start(x)
round(par0, 3)
```

`default_start()`: $p = 0.5$, quartiles for $\mu$, $\text{sd}(x)/2$ for $\sigma$.
:::

::: {.column width="55%"}
```{r}
#| label: fig-data
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.3
grid <- seq(min(x), max(x), length.out = 400)
ggplot(data.frame(x), aes(x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "grey85", colour = "white") +
  geom_line(data = data.frame(x = grid, d = dtmix(grid, par_true, nu)), aes(x, d), linewidth = 1) +
  labs(y = "density", title = "N = 400, true density")
```
:::
::::

## Missing data: the component labels

**Idea:** if we knew which component each $x_i$ came from, estimation would be easy — just like the two coins in lecture 9.

Introduce the (unobserved) label $Z_i \in \{1, 2\}$:
$$
P(Z_i = 1) = p, \qquad X_i \mid Z_i = j \sim t(\mu_j, \sigma_j^2, \nu_j).
$$

Write $f_j(x) = f(x \mid \mu_j, \sigma_j^2, \nu_j)$. The joint density of one complete observation $(x_i, z_i)$ is
$$
f(x_i, z_i \mid \theta) = \big(p\, f_1(x_i)\big)^{\mathbf{1}(z_i = 1)} \big((1 - p)\, f_2(x_i)\big)^{\mathbf{1}(z_i = 2)}.
$$

. . .

**Check:** summing over $z_i \in \{1, 2\}$ gives $p f_1(x_i) + (1-p) f_2(x_i)$ — exactly the mixture density. So $(X, Z)$ is a valid complete-data model for our observed $X$.

**Complete-data log-likelihood** (what we could maximize if $Z$ were observed):
$$
\ell_c(\theta) = \sum_{i=1}^N \mathbf{1}(Z_i = 1)\big(\log p + \log f_1(x_i)\big) + \mathbf{1}(Z_i = 2)\big(\log(1-p) + \log f_2(x_i)\big)
$$

With $Z$ known: $p$ = fraction of labels equal to 1, and $(\mu_j, \sigma_j)$ = a $t$-fit to group $j$ alone. But $Z$ is missing.

::: notes
The trick of EM is to choose missing data such that the complete-data problem is easy. Here the missing data are the labels. The check that summing out z gives the mixture density is what makes this a legal choice of complete data.
:::

## E-step: from $\ell_c$ to $Q$

$Q(\theta \mid \theta') = E_{\theta'}\big(\ell_c(\theta) \mid X = x\big)$ — average $\ell_c$ over what the labels could be, given the data and the current guess $\theta'$.

**Step 1 — linearity.** Given $x$, the terms $\log p + \log f_1(x_i)$ are just numbers, so the expectation only hits the indicators:
$$
E_{\theta'}\big(\mathbf{1}(Z_i = 1) \mid X = x\big) = P_{\theta'}(Z_i = 1 \mid X_i = x_i) =: \gamma_i
$$
(the observations are independent, so only $x_i$ matters for $Z_i$).

**Step 2 — Bayes' formula.**
$$
\gamma_i = \frac{P(Z_i = 1)\, f(x_i \mid Z_i = 1)}{f(x_i)} = \frac{p' f_1'(x_i)}{p' f_1'(x_i) + (1 - p') f_2'(x_i)}
$$

**Result:**
$$
Q(\theta \mid \theta') = \sum_{i=1}^N \gamma_i \big(\log p + \log f_1(x_i)\big) + (1 - \gamma_i)\big(\log(1-p) + \log f_2(x_i)\big)
$$

The $\gamma_i$ are computed from the **old** $\theta'$ and are fixed numbers ("soft labels"); $\theta$ in the logs is the **new** parameter we optimize over.

```{r}
#| label: fig-gamma
#| echo: false
#| fig-height: 2.6
par_hat0 <- em(par0, create_em_aug_step(x, nu))
gdf <- rbind(
  data.frame(x = grid, gamma = e_step(par0, grid, nu), theta = "at the start value"),
  data.frame(x = grid, gamma = e_step(par_hat0, grid, nu), theta = "at the estimate")
)
ggplot(gdf, aes(x, gamma, colour = theta)) +
  geom_line(linewidth = 1) +
  geom_rug(data = data.frame(x), aes(x), inherit.aes = FALSE, alpha = 0.3) +
  labs(y = expression(gamma == P(Z == 1 ~ "|" ~ X == x)), colour = NULL)
```

Far to the right $\gamma$ rises again: component 1 has heavier tails ($\nu_1 = 5$, larger $\sigma_1$), so an extreme point is better explained by component 1 than by component 2.

## M-step: maximize $Q$ — what can be solved?

$Q$ splits into three parts that can be maximized separately:
$$
Q = \underbrace{\sum_i \gamma_i \log p + (1 - \gamma_i)\log(1-p)}_{\text{only } p}
  + \underbrace{\sum_i \gamma_i \log f_1(x_i)}_{\text{only } \mu_1, \sigma_1}
  + \underbrace{\sum_i (1 - \gamma_i) \log f_2(x_i)}_{\text{only } \mu_2, \sigma_2}
$$

**$p$:** $\ \partial_p Q = \frac{\sum_i \gamma_i}{p} - \frac{N - \sum_i \gamma_i}{1 - p} = 0 \;\Rightarrow\; p = \frac{1}{N}\sum_i \gamma_i$ — closed form.

. . .

**$(\mu_1, \sigma_1)$:** a *weighted* $t$ log-likelihood. With $r = x - \mu_1$ and $\log f_1(x) = c(\nu_1) - \log \sigma_1 - \frac{\nu_1 + 1}{2}\log\big(1 + \frac{r^2}{\nu_1 \sigma_1^2}\big)$, the chain rule gives
$$
\partial_{\mu_1} \log f_1(x) = \frac{\nu_1 + 1}{2}\cdot\frac{2r / (\nu_1\sigma_1^2)}{1 + r^2/(\nu_1\sigma_1^2)} = \frac{w\, r}{\sigma_1^2},
\qquad
\partial_{\sigma_1} \log f_1(x) = -\frac{1}{\sigma_1} + \frac{w\, r^2}{\sigma_1^3},
\qquad w = \frac{\nu_1 + 1}{\nu_1 + r^2/\sigma_1^2}.
$$

Summing with weights $\gamma_i$ gives the gradient of $Q$:
$$
\partial_{\mu_1} Q = \sum_i \gamma_i \frac{w_{i1} r_{i1}}{\sigma_1^2}, \qquad
\partial_{\sigma_1} Q = \sum_i \frac{\gamma_i}{\sigma_1}\left(\frac{w_{i1} r_{i1}^2}{\sigma_1^2} - 1\right) \qquad (\text{same for component 2 with } 1 - \gamma_i).
$$

. . .

Setting $\partial_{\mu_1} Q = 0$ gives $\mu_1 = \frac{\sum_i \gamma_i w_{i1} x_i}{\sum_i \gamma_i w_{i1}}$ — it *looks* like a weighted mean, but $w_{i1}$ depends on $\mu_1$ and $\sigma_1$ themselves. It is a fixed-point equation, not a formula. (For a Gaussian mixture $w_{i1} \equiv 1$ and it *is* a formula.)

[No closed-form M-step for $(\mu, \sigma)$. **Generalized EM:** we only need $Q(\theta_{n+1} \mid \theta_n) \ge Q(\theta_n \mid \theta_n)$, so: exact update of $p$ + one uphill (gradient) step for $(\mu_1, \mu_2, \sigma_1, \sigma_2)$.]{.problem}

## From formulas to code: the E-step

Numerical detail: for an outlier, $f_1(x_i)$ and $f_2(x_i)$ can both be tiny, and $\gamma_i$ is a ratio of two tiny numbers. So everything is computed on **log scale**:
$$
a_i = \log p + \log f_1(x_i), \quad b_i = \log(1-p) + \log f_2(x_i), \quad
\gamma_i = \frac{e^{a_i}}{e^{a_i} + e^{b_i}} = \frac{1}{1 + e^{b_i - a_i}} = \texttt{plogis}(a_i - b_i)
$$
and $\ell(\theta) = \sum_i \log(e^{a_i} + e^{b_i})$ with the log-sum-exp trick.

::: {.small-code}
```{r}
#| eval: false
#| code: !expr c(show_fun("ldt_ls"), "", show_fun("log_comp"), "", show_fun("e_step"), "", show_fun("loglik"))
```
:::

## From formulas to code: $Q$ and its gradient

| Math | Code |
|------|------|
| $\gamma_i$ (from old $\theta'$) | `gam = e_step(par0, x, nu)` — can be passed in, so the E-step is done *once* per iteration even if $Q$ is evaluated many times in the line search |
| $\log p + \log f_1(x_i)$ | `log(par[1]) + ldt_ls(x, par[2], par[4], nu[1])` |
| $r_{i1}$, $w_{i1}$ | `r1 <- x - par[2]`, `w1 <- (nu[1] + 1) / (nu[1] + r1^2 / s1^2)` |
| $\partial_{\mu_1} Q = \sum_i \gamma_i w_{i1} r_{i1} / \sigma_1^2$ | `sum(gam * w1 * r1) / s1^2` |

::: {.small-code}
```{r}
#| eval: false
#| code: !expr c(show_fun("Q"), "", show_fun("grad_Q"))
```
:::

## Checking the derivation numerically

Before building algorithms on top, check each formula against something independent:

```{r}
#| label: check-derivation
par_test <- c(0.4, -0.5, 2.5, 1.3, 0.9)

# E-step on log scale vs. the textbook formula with dt()
f1 <- par0[1] * dt((x - par0[2]) / par0[4], nu[1]) / par0[4]
f2 <- (1 - par0[1]) * dt((x - par0[3]) / par0[5], nu[2]) / par0[5]
all.equal(e_step(par0, x, nu), f1 / (f1 + f2))

# analytic gradient of Q vs. numerical differentiation
g <- rbind(analytic  = grad_Q(par_test, par0, x, nu),
           numerical = numDeriv::grad(function(p) Q(p, par0, x, nu), par_test))
colnames(g) <- par_names
g

# Fisher's identity: grad_Q(theta | theta) is the gradient of the log-likelihood
g <- rbind(fisher_identity = grad_loglik(par_test, x, nu),
           numerical       = numDeriv::grad(function(p) loglik(p, x, nu), par_test))
colnames(g) <- par_names
g
```

[Deriving by hand and checking against `numDeriv` catches sign and chain-rule errors before they turn into a mysteriously slow or wrong algorithm.]{.lesson}

# Version 1 → version 2

## GEM, version 1: one gradient step with a fixed step size

```{r}
#| eval: false
#| code: !expr show_fun("create_gem_step_v1")
```

```{r}
#| label: fig-v1
#| echo: false
#| fig-height: 3.6
v1_traces <- do.call(rbind, lapply(c(1e-2, 1e-3, 1e-4), function(g) {
  tr <- trace_run(function(cb) em(par0, create_gem_step_v1(x, nu, g), maxit = 150, cb = cb), x, nu, "")
  data.frame(n = tr$n, loglik = tr$loglik, gamma = format(g))
}))
ggplot(v1_traces, aes(n, loglik, colour = gamma)) +
  geom_line(linewidth = 0.9) +
  coord_cartesian(ylim = c(-800, -783)) +
  labs(x = "iteration", y = expression(l(theta[n])), colour = "step size")
```

[With step size 0.01 the log-likelihood **decreases** in many iterations — the step does not increase $Q$, so it is not a GEM algorithm. With $10^{-4}$ it crawls. The right step size depends on $N$ and $\sigma^2$ (the gradient scales like $N/\sigma^2$).]{.problem}

## GEM, version 2: backtracking line search on $Q$

A GEM step only has to satisfy $Q(\theta_{n+1} \mid \theta_n) \ge Q(\theta_n \mid \theta_n)$. Enforce it with an Armijo condition:

::: {.small-code}
```{r}
#| eval: false
#| code: !expr show_fun("create_gem_step")
```
:::

[A GEM needs a *guaranteed* increase of $Q$ — that is exactly what backtracking gives. Then the ascent property $\ell(\theta_{n+1}) \ge \ell(\theta_n)$ follows from the EM theory.]{.lesson}

::: notes
The warm argument was added later, after profiling. Start by explaining warm = FALSE.
:::

## Running it

```{r}
#| label: run-gem
theta_gem <- em(par0, create_gem_step(x, nu))
theta_gem
loglik(theta_gem, x, nu)
```

`em()` is the driver from the lecture (same stopping rule: $\|\theta_n - \theta_{n-1}\|^2 \le \varepsilon(\|\theta_n\|^2 + \varepsilon)$), extended so the criterion is a function argument. More on that later.

# Testing

## How can I test that the implementation is correct?

| What | How |
|------|-----|
| Building blocks | `ldt_ls()` against `dt()`; mixture density integrates to 1 |
| Analytic gradient of $Q$ | against `numDeriv::grad()` |
| Fisher's identity $\nabla \ell(\theta) = \nabla_\theta Q(\theta \mid \theta)$ | against `numDeriv::grad(loglik)` |
| Different implementations of the same step | naive, readable, optimized R and Rcpp give the same result |
| GEM property | $Q(\theta_{n+1}\mid\theta_n) \ge Q(\theta_n\mid\theta_n)$ and $\ell(\theta_n)$ non-decreasing |
| Convergence | all 7 methods reach the same point; $\nabla\ell(\hat\theta) \approx 0$; $\hat\theta$ is a fixed point of the EM map |
| Statistical sanity | large $N$ recovers the true parameters |
| Fisher information | three different methods agree |

```{r}
#| eval: false
test_that("gradient of Q matches numerical differentiation", {
  num <- numDeriv::grad(function(p) Q(p, par0, x, nu), par_rand)
  expect_equal(grad_Q(par_rand, par0, x, nu), num, tolerance = 1e-6)
})
```

## Running the test suite

```{r}
#| label: run-tests
test_res <- testthat::test_file("tests/test-tmix.R", reporter = "silent")
as.data.frame(test_res)[, c("test", "nb", "failed", "error")]
```

. . .

[While writing the tests I passed the data to `numDeriv::jacobian(f, par, x = x)` — but `x` is also the name of `jacobian()`'s own argument, so it silently differentiated with respect to the *data*. Now data are passed via closures.]{.lesson}

# Alternative solutions

## Idea: add *more* missing data

A $t$-variable is a scale mixture of normals: with $W \mid Z = j \sim \Gamma(\nu_j/2, \text{rate} = \nu_j/2)$ and
$$
X \mid W = w, Z = j \sim \mathcal{N}(\mu_j, \sigma_j^2 / w)
$$
we get $X \mid Z = j \sim t(\mu_j, \sigma_j^2, \nu_j)$. Complete data $(X, Z, W)$:

- $W \mid X = x, Z = j \sim \Gamma\big(\tfrac{\nu_j + 1}{2}, \tfrac{\nu_j + (x-\mu_j)^2/\sigma_j^2}{2}\big)$, so $E(W \mid x, Z = j) = \frac{\nu_j+1}{\nu_j + (x-\mu_j)^2/\sigma_j^2} = w_{ij}$
- $Q_{\text{aug}}$ is a *weighted Gaussian* log-likelihood → **closed-form M-step**:

$$
p = \tfrac1N \textstyle\sum_i \gamma_i, \qquad
\mu_1 = \frac{\sum_i \gamma_i w_{i1} x_i}{\sum_i \gamma_i w_{i1}}, \qquad
\sigma_1^2 = \frac{\sum_i \gamma_i w_{i1} (x_i - \mu_1)^2}{\sum_i \gamma_i}
$$

::: {.small-code}
```{r}
#| eval: false
#| code: !expr c(show_fun("m_step_aug"), "", show_fun("create_em_aug_step"))
```
:::

## Insight: the augmented EM is a *scaled* gradient step on $Q$

Using $\partial_{\mu_1} Q = \sum_i \gamma_i w_{i1}(x_i - \mu_1)/\sigma_1^2$:

$$
\mu_1^{\text{new}} = \frac{\sum_i \gamma_i w_{i1} x_i}{\sum_i \gamma_i w_{i1}}
= \mu_1 + \frac{\sigma_1^2}{\sum_i \gamma_i w_{i1}}\, \partial_{\mu_1} Q(\theta \mid \theta)
$$

- The augmented EM is **also a GEM for the original $Q$**: it takes a gradient step with an automatically chosen step size $\sigma^2 / \sum_i \gamma_i w_{i1}$, which is guaranteed to increase $Q$ (it is an MM-step).
- Compare with version 1: the step size $\approx \sigma^2 / N_1$ explains why a fixed step size could not work.

. . .

[Two different sets of missing data give two different EM algorithms for the same likelihood. Choosing the missing data cleverly gives a closed-form M-step.]{.lesson}

## More alternatives

**Exact EM** (only $Z$ missing): solve the M-step numerically — a weighted $t$-MLE per component with `optim(method = "BFGS")` in $(\mu, \log\sigma)$.

**Direct optimization of $\ell(\theta)$** — the gradient comes for free from Fisher's identity: `grad_loglik <- function(par, x, nu) grad_Q(par, par, x, nu)`

- Gradient ascent with backtracking on $\ell$ (`create_gd_step()`)
- BFGS via `optim()` in $\eta = (\operatorname{logit} p, \mu_1, \mu_2, \log\sigma_1, \log\sigma_2)$ (`fit_bfgs()`)

. . .

[The GEM gradient step at $\theta_n$ uses $\nabla_\theta Q(\theta_n \mid \theta_n) = \nabla \ell(\theta_n)$ — the *same direction* as gradient ascent on $\ell$. The differences are the exact $p$-update and the line search on $Q$ instead of $\ell$.]{.lesson}

## All seven methods agree

```{r}
#| label: all-methods
fits <- lapply(tmix_methods, function(m) tmix(x, nu, par0, method = m))
data.frame(
  method = tmix_methods,
  iterations = sapply(fits, `[[`, "iter"),
  loglik = sapply(fits, `[[`, "loglik"),
  t(sapply(fits, coef))
)
```

(`em_aug_fast` and `em_aug_cpp` are optimized versions of `em_aug` — see profiling. BFGS "iterations" are gradient evaluations.)

# Code structure

## One driver, many step functions

```{r}
#| eval: false
#| code: !expr c(show_fun("em"), "", show_fun("par_conv"))
```

- Every algorithm (EM, GEM, gradient ascent) is a map $\theta \mapsto \theta_{\text{new}}$ created by a **function factory** (`create_*_step(x, nu)`), which stores data and state (e.g. the last step size) in its closure.
- The convergence criterion is a **function argument** → *exactly* the same criterion can be used for all algorithms.
- Callback `cb` for tracing (own `make_tracer()`, like `CSwR::tracer()`, which subtracts its own overhead from the time).

## An S3 class on top

```{r}
#| label: s3
fit <- tmix(x, nu, method = "em_aug")
fit
```

:::: {.columns}
::: {.column width="50%"}
```{r}
#| label: s3-plot
#| fig-width: 6
#| fig-height: 3.6
plot(fit)
```
:::
::: {.column width="50%"}
Methods: `print()`, `plot()`, `coef()`, `logLik()` (so `AIC()` works), `vcov()` (Fisher information) and `predict()` (responsibilities).

Files:

- `R/tmix-model.R` — density, $\ell$, E-step, $Q$, $\nabla Q$
- `R/em-algorithms.R` — driver + EM/GEM steps
- `R/direct-methods.R` — gradient ascent, BFGS
- `R/fisher.R`, `R/tracer.R`, `R/tmix-class.R`
- `src/em_aug_step.cpp`, `tests/test-tmix.R`
:::
::::

# Convergence

## Tracing the algorithms

```{r}
#| label: traces
par_hat <- em(par0, create_em_aug_step(x, nu), conv = par_conv(1e-24), maxit = 2000)
ll_hat <- loglik(par_hat, x, nu)
runs <- list(
  em_aug   = function(cb) em(par0, create_em_aug_step(x, nu), cb = cb, conv = par_conv(1e-24), maxit = 150),
  em_exact = function(cb) em(par0, create_em_exact_step(x, nu), cb = cb, conv = par_conv(1e-24), maxit = 150),
  gem      = function(cb) em(par0, create_gem_step(x, nu), cb = cb, conv = par_conv(1e-24), maxit = 150),
  gd       = function(cb) em(par0, create_gd_step(x, nu), cb = cb, conv = par_conv(1e-24), maxit = 150)
)
traces <- do.call(rbind, Map(trace_run, runs, label = names(runs), MoreArgs = list(x = x, nu = nu)))
traces$par_dist <- sqrt(rowSums(sweep(as.matrix(traces[, 3:7]), 2, par_hat)^2))
```

```{r}
#| label: fig-traces
#| echo: false
#| fig-height: 3.8
p1 <- ggplot(traces, aes(n, pmax(ll_hat - loglik, 1e-16), colour = method)) +
  geom_line(linewidth = 0.9) + scale_y_log10() +
  scale_colour_manual(values = method_cols) +
  labs(x = "iteration", y = expression(hat(l) - l(theta[n])))
p2 <- ggplot(traces, aes(n, pmax(par_dist, 1e-16), colour = method)) +
  geom_line(linewidth = 0.9) + scale_y_log10() +
  scale_colour_manual(values = method_cols) +
  labs(x = "iteration", y = expression(group("||", theta[n] - hat(theta), "||")))
p1 + p2 + plot_layout(guides = "collect")
```

All four increase $\ell$ monotonically and converge **linearly** (straight lines on log scale).

## The stopping criterion is not neutral

```{r}
#| label: criterion
crit <- expand.grid(eps = 10^-c(6, 8, 10, 12), method = c("em_aug", "gem"), stringsAsFactors = FALSE)
crit_res <- t(mapply(function(eps, method) {
  step <- if (method == "em_aug") create_em_aug_step(x, nu) else create_gem_step(x, nu)
  th <- em(par0, step, eps = eps)
  c(iter = attr(th, "iter"), gap = ll_hat - loglik(th, x, nu),
    grad_norm = sqrt(sum(grad_loglik(th, x, nu)^2)) / length(x))
}, crit$eps, crit$method))
cbind(crit, signif(crit_res, 3))
```

- Near $\hat\theta$: $\theta_n - \hat\theta \approx \rho\,(\theta_{n-1} - \hat\theta)$, so $\|\theta_n - \hat\theta\| \approx \frac{\rho}{1-\rho}\|\theta_n - \theta_{n-1}\|$.
- A slowly converging algorithm (large $\rho$) takes small steps and **stops further from the optimum** for the same $\varepsilon$.

[Same $\varepsilon$ ≠ same accuracy. For benchmarks I either stop on the **gradient** $\|\nabla\ell(\theta_n)\|/N \le \text{tol}$ (cheap thanks to Fisher's identity: `grad_conv()`) or compare convergence as a function of **real time**.]{.lesson}

# Benchmarking

## `bench::mark` with the same (gradient) criterion

```{r}
#| label: bench-methods
conv <- grad_conv(x, nu, tol = 1e-7)
same_estimate <- function(a, b) isTRUE(all.equal(as.vector(a), as.vector(b), tolerance = 1e-5))
bm <- bench::mark(
  em_aug   = em(par0, create_em_aug_step(x, nu), conv = conv),
  em_exact = em(par0, create_em_exact_step(x, nu), conv = conv),
  gem      = em(par0, create_gem_step(x, nu, warm = TRUE), conv = conv),
  gem_cold = em(par0, create_gem_step(x, nu), conv = conv),
  gd       = em(par0, create_gd_step(x, nu), conv = conv),
  bfgs     = fit_bfgs(par0, x, nu, reltol = 1e-12),
  check = same_estimate, min_iterations = 20 # all methods must give the same estimate
)
res <- bm$result
data.frame(method = as.character(bm$expression), median = format(bm$median), mem = format(bm$mem_alloc),
  iter = sapply(res, attr, "iter"),
  gap = signif(sapply(res, function(p) ll_hat - loglik(p, x, nu)), 2))
```

```{r}
#| label: fig-bench
#| echo: false
#| fig-height: 3
plot(bm, type = "violin") + labs(x = NULL)
```

## Convergence as a function of real time

A larger data set ($N = 5000$) so that times are not dominated by overhead:

```{r}
#| label: time-traces
#| code-fold: true
set.seed(7)
x_big <- rtmix(5000, par_true, nu)
p0_big <- default_start(x_big)
ll_hat_big <- loglik(em(p0_big, create_em_aug_step(x_big, nu), conv = par_conv(1e-24), maxit = 3000), x_big, nu)
runs_big <- list(
  em_aug   = function(cb) em(p0_big, create_em_aug_step(x_big, nu), cb = cb, conv = par_conv(1e-24), maxit = 200),
  em_exact = function(cb) em(p0_big, create_em_exact_step(x_big, nu), cb = cb, conv = par_conv(1e-24), maxit = 60),
  gem      = function(cb) em(p0_big, create_gem_step(x_big, nu, warm = TRUE), cb = cb, conv = par_conv(1e-24), maxit = 200),
  gem_cold = function(cb) em(p0_big, create_gem_step(x_big, nu), cb = cb, conv = par_conv(1e-24), maxit = 200),
  gd       = function(cb) em(p0_big, create_gd_step(x_big, nu), cb = cb, conv = par_conv(1e-24), maxit = 300),
  bfgs     = function(cb) fit_bfgs(p0_big, x_big, nu, reltol = 1e-16, cb = cb)
)
time_traces <- do.call(rbind, Map(trace_run, runs_big, label = names(runs_big),
                                  MoreArgs = list(x = x_big, nu = nu)))
# BFGS is traced in every function evaluation (also in line searches): use the best value so far
time_traces <- do.call(rbind, lapply(split(time_traces, time_traces$method),
  function(d) transform(d, gap = pmax(ll_hat_big - cummax(loglik), 1e-12))))
```

```{r}
#| label: fig-time
#| echo: false
#| fig-height: 3.8
ggplot(time_traces, aes(1000 * time, gap, colour = method)) +
  geom_line(linewidth = 0.9) + scale_y_log10() +
  scale_colour_manual(values = method_cols) +
  coord_cartesian(xlim = c(0, 400)) +
  labs(x = "time (ms)", y = expression(hat(l) - l(theta[n])))
```

[Per iteration the exact EM is best, but its M-step is expensive. In real time the cheap closed-form augmented EM and BFGS win. The GEM trades fewer iterations for line-search cost.]{.lesson}

# Profiling

## Profiling the GEM (version 2)

```{r}
#| label: prof-gem
prof_file <- tempfile()
Rprof(prof_file, interval = 0.005)
for (i in 1:15) em(p0_big, create_gem_step(x_big, nu))
Rprof(NULL)
prof <- summaryRprof(prof_file)$by.total
prof[rownames(prof) %in% dQuote(c("em", "em_step", "Q", "ldt_ls", "grad_Q", "e_step", "plogis"), FALSE), ]
```

`profvis::profvis(em(p0_big, create_gem_step(x_big, nu)))` gives the same picture interactively (see `R/profiling-demo.R`).

. . .

[Almost all time is spent in `Q()` → `ldt_ls()` **inside the line search**: backtracking starts from $\gamma_0 = 1$ in every iteration and needs ~10 halvings before $\gamma \approx \sigma^2/N$.]{.problem}

## Fix: warm-start the line search

Start from the last accepted step size divided by $d$ (i.e. try a slightly *larger* step first): `warm = TRUE`.

```{r}
#| label: bench-warm
bm_warm <- bench::mark(
  cold = em(p0_big, create_gem_step(x_big, nu, warm = FALSE), conv = grad_conv(x_big, nu, 1e-7)),
  warm = em(p0_big, create_gem_step(x_big, nu, warm = TRUE), conv = grad_conv(x_big, nu, 1e-7)),
  check = same_estimate, min_iterations = 5
)
data.frame(expression = as.character(bm_warm$expression), median = format(bm_warm$median),
           mem_alloc = format(bm_warm$mem_alloc), iter = sapply(bm_warm$result, attr, "iter"))
```

[Profiling pointed to a single line of code. Keeping *state* (the step size) in the closure of the step function halved the run time.]{.lesson}

## Profiling the augmented EM

```{r}
#| label: prof-aug
Rprof(prof_file, interval = 0.005)
for (i in 1:15) em(p0_big, create_em_aug_step(x_big, nu), maxit = 100)
Rprof(NULL)
head(summaryRprof(prof_file)$by.self, 6)
```

Hot spots: `plogis()`, `ldt_ls()` (recomputes `lgamma()` constants every time), `cbind()` in `log_comp()`, and the squared residuals are computed twice (in `e_step()` and for $w_{ij}$).

## My first "optimization" made things slower

My first idea was to avoid `exp(log1p(.))` by computing $(1 + d/\nu)^{-(\nu+1)/2}$ directly. That version was *slower*. Measuring the pieces explains why:

```{r}
#| label: micro
r <- rnorm(5000)
bench::mark(
  pow_integer    = (1 + r^2 / 5)^-3,      # nu = 5:  exponent -3
  pow_fractional = (1 + r^2 / 10)^-5.5,   # nu = 10: exponent -5.5
  exp_log1p      = exp(-5.5 * log1p(r^2 / 10)),
  plogis         = plogis(r),
  one_over       = 1 / (1 + exp(-r)),
  dt             = dt(r, 10),
  check = FALSE
)[, c("expression", "median", "itr/sec")]
```

[`^` is only fast for *integer* exponents (R special-cases them); with $\nu_2 = 10$ the exponent is $-5.5$ and `pow()` is slower than `exp(log1p())`. `dt()` is slow, and `plogis()` is slower than `1/(1 + exp(-x))`. Measure, don't guess.]{.lesson}

## Optimized R and Rcpp

:::: {.columns}
::: {.column width="50%"}
::: {.small-code}
```{r}
#| eval: false
#| code: !expr show_fun("create_em_aug_step_fast")
```
:::
:::
::: {.column width="50%"}
::: {.small-code}
```{r}
#| echo: false
#| class-output: cpp
cat(readLines("src/em_aug_step.cpp")[25:50], sep = "\n")
```
:::
The C++ version does E- and M-step in two passes over the data, without allocating any R vectors.
:::
::::

## `bench::press`: cost per EM step as a function of $N$

```{r}
#| label: press
press_res <- bench::press(N = c(500, 2000, 8000, 32000), {
  xN <- rtmix(N, par_true, nu)
  pN <- default_start(xN)
  s_v1 <- create_em_aug_step_v1(xN, nu)
  s_readable <- create_em_aug_step(xN, nu)
  s_fast <- create_em_aug_step_fast(xN, nu)
  bench::mark(v1 = s_v1(pN), readable = s_readable(pN), fast = s_fast(pN),
              cpp = em_aug_step_cpp(xN, pN, nu), check = TRUE) # check = TRUE: same result
})
```

```{r}
#| label: fig-press
#| echo: false
#| fig-height: 3.3
press_df <- data.frame(expression = as.character(press_res$expression), N = press_res$N,
                       median = as.numeric(press_res$median), mem = as.numeric(press_res$mem_alloc))
q1 <- ggplot(press_df, aes(N, 1e6 * median, colour = expression)) + geom_line() + geom_point() +
  scale_x_log10() + scale_y_log10() + labs(y = "median time (µs)", colour = NULL)
v1_time <- with(subset(press_df, expression == "v1"), setNames(median, N))
press_df$speedup <- v1_time[as.character(press_df$N)] / press_df$median
q2 <- ggplot(press_df, aes(N, speedup, colour = expression)) +
  geom_line() + geom_point() + scale_x_log10() + labs(y = "speed-up relative to v1", colour = NULL)
q1 + q2 + plot_layout(guides = "collect")
```

[The biggest gain came from dropping `dt()` and the data frame (v1 → readable). Rcpp only gives a modest extra factor: vectorized R already runs in C, and the time is dominated by `log1p()` and `exp()`, which cost the same in C++. Rcpp's real win here is **zero memory allocation**.]{.lesson}

# Theory check

## Fisher information — three ways (lecture 10)

```{r}
#| label: fisher
i_score <- fisher_score(par_hat, x, nu)          # 1. differentiate the score (Fisher's identity)
louis   <- fisher_louis(par_hat, x, nu)          # 2. i_X = i_Y - i_{Y|X}, complete data (X, Z)
em_map  <- fisher_em_map(par_hat, x, nu)         # 3. i_X = (I - DPhi^T) i_Y, augmented EM map
c(louis = max(abs(louis$iX - i_score)), em_map = max(abs(em_map$iX - i_score)))
rbind(estimate = par_hat, se = sqrt(diag(solve(i_score))))
```

The augmented EM map uses *other* complete data $(X, Z, W)$, so its $i_Y$ and $D\Phi$ differ from method 2 — but $i_X$ is the same, as it must be.

## Convergence rates: theory vs practice

Rate of the EM algorithm $= \rho(D\Phi(\hat\theta)) = \rho(i_{Y\mid X}\, i_Y^{-1})$ = "fraction of missing information".

```{r}
#| label: rates
rate_emp <- sapply(split(traces, traces$method), function(d) {
  d <- subset(d, par_dist > 1e-6 & par_dist < 1e-2) # linear phase, above the numerical floor
  unname(exp(coef(lm(log(par_dist) ~ n, data = d))["n"]))
})
data.frame(
  method = names(rate_emp),
  empirical = round(rate_emp, 3),
  theory = c(em_aug = spectral_radius(em_map$DPhi),
             em_exact = spectral_radius(louis$iYX %*% solve(louis$iY)),
             gem = NA, gd = NA)[names(rate_emp)]
)
```

- The empirical rates (slope of $\log\|\theta_n - \hat\theta\|$) match the theoretical rates.
- All traces flatten at $\|\theta_n - \hat\theta\| \approx 10^{-8}$: there $\hat\ell - \ell(\theta_n) \approx 10^{-13} \approx$ machine precision $\times |\ell|$, so no algorithm can see further progress. The rates are estimated above this floor.
- The augmented EM treats $W$ as missing too → **more missing information → slower rate** than the exact EM, but each iteration is much cheaper.
- The GEM has no fixed EM map (the step size comes from a line search), so there is no theoretical rate — empirically it is `r round(rate_emp[["gem"]], 2)`, as good as the exact EM, because the line search along $\nabla\ell$ picks near-optimal steps.

[This explains the benchmark: "we trade slower convergence for faster iterations" (lecture 10), and the numbers confirm the implementation of both EM maps.]{.lesson}

## Do the standard errors make sense? 200 simulated data sets

```{r}
#| label: sim-study
set.seed(123)
sims <- replicate(200, {
  xs <- rtmix(400, par_true, nu)
  th <- em(par_true, function(p) em_aug_step_cpp(xs, p, nu), eps = 1e-14)
  c(th, sqrt(diag(solve(fisher_score(th, xs, nu)))))
})
data.frame(true = par_true, mean_estimate = rowMeans(sims[1:5, ]),
           empirical_sd = apply(sims[1:5, ], 1, sd), mean_fisher_se = rowMeans(sims[6:10, ]))
```

The Fisher-information standard errors agree with the actual sampling variability (started at the truth to stay at the "correct" labeling — see next slide).

# Robustness

## Starting values: 100 random starts

```{r}
#| label: multistart
set.seed(1)
starts <- replicate(100, c(runif(1, 0.1, 0.9), runif(2, min(x), max(x)), runif(2, 0.2, 3)))
ms <- do.call(rbind, lapply(c("em_aug", "gem", "bfgs"), function(m) {
  do.call(rbind, lapply(seq_len(ncol(starts)), function(i) {
    f <- tmix(x, nu, starts[, i], method = m)
    data.frame(method = m, t(coef(f)), loglik = f$loglik)
  }))
}))
ms$solution <- with(ms, ifelse(pmin(p, 1 - p) < 1e-3, "a component vanishes (p -> 0 or 1)",
                        ifelse(pmin(sigma1, sigma2) < 0.1, "spike (sigma -> 0)",
                        ifelse(mu1 > mu2, "labels swapped", "labels as simulated"))))
table(ms$solution, ms$method)
aggregate(cbind(loglik, p, mu1, mu2, sigma1, sigma2) ~ solution, ms, median)
```

## What the multi-start taught me

- With $\nu_1 \neq \nu_2$, swapping the labels gives a **different** model: here the $t_5$-component on the right cluster fits *better* ($\ell = -782.2$ vs $-784.3$). Two genuine local maxima → always compare several starts (at least both labelings).
- The **augmented EM always found one of the two maxima**. GEM and BFGS sometimes let a component vanish ($p \to 0$): $p$ collapses faster than the gradient steps can move $(\mu, \sigma)$ to the data, and $p = 0$ is absorbing for EM ($\gamma_i = 0$ for all $i$).
- BFGS can also run into the unbounded-likelihood "spike" ($\sigma_j \to 0$ on a single point).

[Robustness towards starting values is a property of the algorithm, not only of the likelihood. The closed-form augmented EM was both the fastest and the most robust.]{.lesson}

## Outliers: Gaussian vs $t$ mixture

20 data sets ($N = 400$); add $k$ outliers $\pm U(8, 20)$; fit both mixtures by EM (started at the truth).

```{r}
#| label: outliers
#| code-fold: true
set.seed(3)
out_res <- do.call(rbind, lapply(1:20, function(rep) {
  x0 <- rtmix(400, par_true, nu)
  out <- runif(40, 8, 20) * sample(c(-1, 1), 40, replace = TRUE)
  do.call(rbind, lapply(c(0, 5, 10, 20, 40), function(k) {
    xc <- c(x0, out[seq_len(k)])
    rbind(
      data.frame(k, model = "t mixture", t(em(par_true, create_em_aug_step(xc, nu)))),
      data.frame(k, model = "Gaussian mixture", t(em(par_true, create_em_gauss_step(xc), maxit = 5000)))
    )
  }))
}))
out_long <- reshape(out_res, direction = "long", varying = par_names, v.names = "estimate",
                    timevar = "parameter", times = par_names)
```

```{r}
#| label: fig-outliers
#| echo: false
#| fig-height: 3.6
out_med <- aggregate(estimate ~ k + model + parameter, out_long, median)
ggplot(out_med, aes(k, estimate, colour = model)) +
  geom_hline(data = data.frame(parameter = par_names, value = par_true), aes(yintercept = value), linetype = 2) +
  geom_line(linewidth = 0.9) + geom_point() +
  facet_wrap(~parameter, nrow = 1, scales = "free_y") +
  labs(x = "number of outliers", y = "median estimate", colour = NULL) +
  theme(legend.position = "bottom")
```

[Already 5 outliers (1%) ruin the Gaussian mixture's $\sigma_1$, and at 20 outliers one Gaussian component is used for the outliers alone. The $t$-mixture degrades gracefully — but with fixed $\nu$ it is not immune either.]{.lesson}

# Summary

## What I learned

1. **Correctness**: a GEM step must *guarantee* an increase of $Q$ (v1 did not). Testing with `numDeriv`, Fisher's identity and monotonicity caught the bugs — including a silent argument-name clash in `numDeriv`.
2. **Alternatives**: choosing more missing data ($W$) gives a closed-form EM, which is also a scaled gradient step on $Q$. Exact EM, gradient ascent and BFGS all reach the same point.
3. **Structure**: one generic driver + step-function factories + pluggable convergence criterion; an S3 class `tmix` on top.
4. **Benchmarking**: the same $\varepsilon$ does not mean the same accuracy — compare with a gradient criterion or against real time.
5. **Profiling**: the line search dominated the GEM (fixed by a warm start, 2× faster); `dt()`/`plogis()` dominated the EM. My own guess (`pow`) was wrong. Rcpp gave a modest gain and zero allocations.
6. **Theory matches practice**: three Fisher information methods agree, SEs match simulations, and the observed convergence rates equal $\rho(D\Phi)$.
7. **Robustness**: multiple local maxima (labelings), vanishing components for gradient-based methods, and the $t$-mixture is far more robust to outliers than the Gaussian mixture.
